% -*-latex-*-
\nusmv is a symbolic model checker originated from the
reengineering, reimplementation and extension of CMU \smv, the
original BDD-based model checker developed at CMU \cite{McMil93}.
The \nusmv project aims at the development of a state-of-the-art
symbolic model checker, designed to be applicable in technology
transfer projects: it is a well structured, open, flexible and
documented platform for model checking, and is robust and close to
industrial systems standards \cite{CCGR00}.

Version 1 of \nusmv basically implements BDD-based symbolic model checking.
Version 2 of \nusmv (\nusmvTwo in the following) 
inherits all the functionalities of the previous version, and extends them in
several directions \cite{CCG+02}.
The main novelty in \nusmvTwo is the integration of model checking
techniques based on propositional satisfiability (SAT) \cite{BCCZ99}.
SAT-based model checking is currently enjoying a substantial success
in several industrial fields, and opens up new research directions.
BDD-based and SAT-based model checking are often able to solve
different classes of problems, and can therefore be seen as
complementary techniques.

Starting from \nusmvTwo, we are also adopting a new development and
license model. \nusmvTwo is distributed with an OpenSource
license\footnote{(see \url{http://www.opensource.org})}, that
allows anyone interested to freely use the tool and to participate in
its development. The aim of the \nusmv OpenSource project is to
provide to the model checking community a common platform for the
research, the implementation, and the comparison of new symbolic model
checking techniques. Since the release of \nusmvTwo, the \nusmv team
has received code contributions for different parts of the system.
Several research institutes and commercial companies have expressed
interest in collaborating to the development of \nusmv. The main
features of \nusmv are the following:
\begin{itemize}
\item {\bf Functionalities.}
\nusmv allows for the representation of synchronous and
asynchronous finite state systems\footnote{However, asynchronous
processes are deprecated in version 2.5.0 and later, and may be no
longer supported in future versions.}, and for the analysis of
specifications expressed in Computation Tree Logic (CTL) and Linear
Temporal Logic (LTL), using BDD-based and SAT-based model checking
techniques.  Heuristics are available for achieving efficiency and
partially controlling the state explosion.  The interaction with
the user can be carried on with a textual interface, as well as in
batch mode.
\item{\bf Architecture.}
A software architecture has been defined. The different components and
functionalities of \nusmv have been isolated and separated in
modules. Interfaces between modules have been provided. This 
reduces the effort needed to modify and extend \nusmv.
\item{\bf Quality of the implementation.}
\nusmv is written in ANSI C, is POSIX compliant, and has been
debugged with Purify in order to detect memory leaks. Furthermore, the
system code is thoroughly commented. \nusmv uses the state of the
art BDD package developed at Colorado University, and provides a general
interface for linking with state-of the-art SAT solvers. This makes
\nusmv very robust, portable, efficient, and easy to
understand by people other than the developers.
\end{itemize}
This document is structured as follows.
\begin{itemize}
%\item In \cref{Tutorial} 
%we give an introduction to the usage of the main 
%functionalities of \nusmv. 
\item In \cref{Input Language} we define the syntax of the input language of
\nusmv.
\item In \cref{Running NuSMV interactively} the commands of the
interaction shell are described.
\item In \cref{Running NuSMV batch}
we define the batch mode of \nusmv.
\end{itemize}
\nusmv is available at \url{http://nusmv.fbk.eu}.
