% -*-latex-*-
\chapter{Compatibility with CMU \smv}
%
The \nusmv language is mostly source compatible with the original
version of \smv distributed at Carnegie Mellon University from which
we started.
%
In this appendix we describe the most common problems that can be
encountered when trying to use old CMU \smv programs with \nusmv.

The main problem is variable names in old programs that conflicts with
new reserved keywords.  
%
The list of the new reserved keywords of \nusmv w.r.t. CMU \smv is the
following:

\begin{table}[h]
\begin{tabular}{p{100pt}p{300pt}}
%
\code{F, G, X, U, V, W, H, O, Y, Z, S, T, B} &
These names are reserved for the LTL temporal operators.\\
%
\code{CTLSPEC} &
It is used to introduce CTL specifications. \\
%
\code{LTLSPEC} &
It is used to introduce LTL specifications. \\
%
\code{INVARSPEC} &
It is used to introduce invariant specifications.\\
%
\code{PSLSPEC} &
It is used to introduce PSL specifications.\\
%
\code{IVAR} &
It is used to introduce input variables. \\
%
\code{FROZENVAR} &
It is used to introduce frozen variables. \\
%
\code{JUSTICE} &
It is used to introduce ``justice'' fairness constraints.\\
%
\code{COMPASSION} &
It is used to introduce ``compassion'' fairness constraints. \\
%
\code{CONSTANTS} &
It is used to force declaration of constants. \\
%
\code{word} &
It is used to declare word type variables. \\
%
\code{word1} &
It is used to cast boolean expressions to word type.\\
%
\code{bool} &
It is used to cast word1 expressions to boolean type.\\
%
\code{unsigned} &
It is used to cast signed words to unsigned ones.\\
%
\code{signed} &
It is used to cast unsigned words to signed ones.\\
%
\code{extend} &
It is used to increase the width of words.\\
%
\end{tabular}
\end{table}

The \code{IMPLEMENTS}, \code{INPUT}, \code{OUTPUT} statements are not
no longer supported by \nusmv.

\nusmv differs from CMU \smv also in the controls that are performed
on the input formulas. 
%
Several formulas that are valid for CMU \smv, but that have no clear
semantics, are not accepted by \nusmv. 

In particular:

\begin{itemize}

\item It is no longer possible to write formulas containing
      nested`\code{next}'.
\begin{alltt}
TRANS
  next(alpha & next(beta | next(gamma))) -> delta
\end{alltt}

\item It is no longer possible to write formulas containing
      `\code{next}' in the right hand side of ``normal'' and ``init''
      assignments (they are allowed in the right hand side of ``next''
      assignments), and with the statements `\code{INVAR}' and
      `\code{INIT}'.

\begin{nusmvCode}
INVAR
  next(alpha) & beta
INIT
  next(beta) -> alpha
ASSIGN
  delta := alpha & next(gamma);       -- normal assignments
  init(gamma) := alpha & next(delta); -- init assignments
\end{nusmvCode}

\item It is no longer possible to write `\code{SPEC}',
      `\code{FAIRNESS}' statements containing `\code{next}'.

\begin{nusmvCode}
FAIRNESS
 next(running)
SPEC
 next(x) & y
\end{nusmvCode}

\item The check for circular dependencies among variables has been 
      done more restrictive.  We say that variable \textit{x} depends on
      variable \textit{y} if \textit{x := f(y)}.  We say that there is a
      circular dependency in the definition of \textit{x} if:

\begin{itemize}
  \item  \textit{x} depends on itself ( e.g. \textit{x := f(x,y)} );
  \item \textit{x} depends on \textit{y} and \textit{y} depends on
        \textit{x} (e.g. \textit{x := f(y)} and \textit{y := f(x)} or
        \textit{x := f(z)}, \textit{z := f(y)} and \textit{y := f(x)} ).
\end{itemize}
%
In the case of circular dependencies among variables there is no fixed
order in which we can compute the involved variables. Avoiding
circular dependencies among variables guarantee that there exists an
order in which the variables can be computed. In \nusmv circular
dependencies are not allowed.

In CMU \smv the test for circular dependencies is able to detect
circular dependencies only in ``normal'' assignments, and not in
``next'' assignments. The circular dependencies check of \nusmv has
been extended to detect circularities also in ``next''
assignments. For instance the following fragment of code is accepted
by CMU \smv but discarded by \nusmv.

\begin{nusmvCode}
MODULE main
VAR
  y : boolean;
  x : boolean;
ASSIGN
  next(x) := x & next(y);
  next(y) := y & next(x);
\end{nusmvCode}
\end{itemize}

Another difference between \nusmv and CMU \smv is in the variable
order file.  
%
The variable ordering file accepted by \nusmv can be partial and can
contain variables not declared in the model. Variables listed in the
ordering file but not declared in the model are simply discarded. 
%
The variables declared in the model but not listed in the variable
file provided in input are created at the end of the given ordering
following the default ordering. 
%
All the ordering files generated by CMU \smv are accepted in input
from \nusmv but the ordering files generated by \nusmv may be not
accepted by CMU \smv.  
%
Notice that there is no guarantee that a good ordering for CMU \smv is
also a good ordering for \nusmv.  
%
In the ordering files for \nusmv, identifier
\code{\_process\_selector\_} can be used to control the position of
the variable that encodes process selection. 
%
In CMU \smv it is not possible to control the position of this
variable in the ordering; it is hard-coded at the top of the ordering.
%
A further difference about variable ordering consists in the fact that
in \nusmv it is allowed to specify single bits of scalar variables. In
the example:
\begin{nusmvCode}
VAR x : 0..7;
\end{nusmvCode}

\nusmv will create three variables \code{x.0}, \code{x.1} and
\code{x.2} that can be explicitly mentioned in the variable ordering
file to fine control their ordering. 
