% -*-latex-*-
When the \commandopt{int} option is not specified, \nusmv
runs as a batch program, in the style of \smv, performing (some
of) the steps described in previous section in a fixed sequence.
\begin{alltt}
\shellprompt \shelltext{\nusmvtxt [command line options] {\it input-file}} \ret
\end{alltt}
The program described in {\it input-file} is processed, and the
corresponding finite state machine is built.  Then, if \emph{input-file}
contains formulas to verify, their truth in the specified structure is
evaluated. For each formula which is not true a counterexample is
printed.\\
The batch mode can be controlled with
the following command line options:\\
\begin{alltt}
\nusmv [-h | -help] [-v {\it vl}] [-int] [[-source script_file | -load script_file]]
       [-s] [-old] [-old_div_op] [-smv_old]
       [-disable_syntactic_checks] [-keep_single_value_vars]
       [-disable_daggifier] [-dcx] [-cpp] [-pre {\it pps}] [-ofm {\it
       fm\_file}] [-obm {\it bm\_file}] [-lp] [-n {\it idx}] [-is]
       [-ic] [-ils] [-ips] [-ii] [-ctt] [[-f] [-r]]|[-df] [-flt] [-AG]
       [-coi] [-i {\it iv\_file}] [-o {\it ov\_file}] [-t {\it
       tv\_file}] [-reorder] [-dynamic] [-m {\it method}]
       [-disable_sexp2bdd_caching] [-bdd_soh heuristics]
       [[-mono]|[-thresh {\it cp\_t}]|[-cp {\it cp\_t}]|[-iwls95 {\it
       cp\_t}]] [-noaffinity] [-iwls95preorder] [-bmc] [-bmc\_length
       {\it k}] [-sat\_solver {\it name}] [-sin on|off] [-rin on|off]
       [-ojeba {\it algorithm}] [{\it input-file}]

\end{alltt}
where the meaning of the options is described below. If
{\it input-file} is not provided in batch mode, then the model is read
from standard input.\\

\begin{nusmvTable}

\opt{-help}{ }

\opt{-h}{%
\index{ \code{-help}}%
\index{ \code{-h}}%
Prints the command line help.}

\opt{-v {\it verbose-level}}{%
\index{\code{-v} {\it verbose-level}}%
Enables printing of additional information on the internal operations of
\nusmv. Setting {\it verbose-level} to \code{1} gives the basic
information. Using this option makes you feel better, since otherwise
the program prints nothing until it finishes, and there is no evidence
that it is doing anything at all. Setting the {\it verbose-level}
higher than 1 enables printing of much extra information.}
\opt{-int} {Enables interactive mode}
\opt{-source {\it sc\_file}}{Executes NuSMV commands from file \filename{sc\_file}}
\opt{-load {\it sc\_file}}{same as -source (deprecated)}
\opt{-s} {%
Avoids to load the \nusmv commands contained in \code{
$\sim$/.nusmvrc} or in \code{.nusmvrc}  or in 
\code{\$\{\stdsyslib\}/master.nusmvrc}.} 
\vindex{\stdsyslib}
\index{master.nusmvrc}
\index{.nusmvrc}
\index{\~/.nusmvrc}

\opt{-old} {%
\index{\code{-old}}%
Keeps backward compatibility with older versions of NuSMV. This option
disables some new features like type checking and dumping of new
extension to SMV files. In addition, if enabled, \code{case}
conditions also accepts ``\code{1}'' which is semantically equivalent
to the truth value ``\code{TRUE}''. This backward compatibility
feature has been added in \nusmv 2.5.1 in order to help porting of old
SMV models. Infact, in versions older than 2.5.1, it was pretty common
to use \code{1} in \code{case} conditions expressions. For an example
please see section \ref{ref::caseconditionexample}}

\opt{-old\_div\_op} {%
\index{\code{-old\_div\_op}}%
Enables the old semantics of ``\code{/}'' and ``\code{mod}'' operations (from
\nusmv 2.3.0) instead of ANSI C semantics.}

\opt{-disable\_syntactic\\\_checks} {%
\index{\code{-disable\_syntactic\_checks}}%
Disables all syntactic checks that will be performed when flattening
the input model. Warning: If the model is not well-formed, \nusmv may
result in unpredictable results, use this option at your own risk.}

\opt{-disable\_daggifier} {%
\index{\code{-disable\_daggifier}}%
Disables the daggification feature of model dumping}

\opt{-keep\_single\\\_value\_vars} {%
\index{\code{-keep\_single\_value\_vars}}%
Does not convert variables that have only one single possible value
into constant DEFINEs}

\opt{-dcx}{%
\index{\code{-dcx}}%
Disables the generation of counter-examples for properties that
are proved to be false. See also variable \envvar{counter\_examples}}

\opt{-cpp}{%
\index{\code{-cpp}}%
Runs pre-processor on \smv files before any of those specified with the
-pre option.}

\opt{-pre {\it pps}}{%
\index{\code{-pre} {\it pps}}%
Specifies a list of pre-processors to run (in the order given) on the
input file before it is parsed by \nusmv. Note that if the
\commandopt{cpp} command is used, then the pre-processors specified by
this command will be run after the input file has been pre-processed
by that pre-processor. {\it pps} is either one single pre-processor
name (with or without double quotes) or it is a space-separated list
of pre-processor names contained within double quotes.}

\opt{-ofm {\it fm\_file}}{%
\index{ \code{-ofm} {\it fm\_file}}%
prints flattened model to file {\it fn\_file}}

\opt{-obm {\it bm\_file}} {%
\index{ \code{-obm} {\it bm\_file}}%
Prints boolean model to file {\it bn\_file}}

\opt{-lp}{%
\index{\code{-lp}}%
Lists all properties in \smv model}

\opt{-n {\it idx}}{%
\index{\code{-n} {\it idx}}%
Specifies which property of \smv model should be checked}

\opt{-is}{%
\index{\code{-is}}%
Does not check \code{SPEC} properties. Sets to ``1''
the \envvar{ignore\_spec} environment variable.}

\opt{-ic}{%
\index{\code{-ic}}%
Does not check \code{COMPUTE} properties. Sets to ``1''
the \envvar{ignore\_compute} environment variable.}

\opt{-ils}{%
\index{\code{-ils}}%
Does not check \code{LTLSPEC} properties. Sets to ``1''
the \envvar{ignore\_ltlspec} environment variable.}

\opt{-ips}{%
\index{\code{-ils}}%
Does not check \code{PSLSPEC} properties. Sets to ``1''
the \envvar{ignore\_pslspec} environment variable.}

\opt{-ii}{%
\index{\code{-ii}}%
Does not check \code{INVARSPEC} properties. Sets to ``1''
the \envvar{ignore\_invariant} environment variable.}

\opt{-ctt}{%
\index{\code{-ctt}}%    
Checks whether the transition relation is total.}

\opt{-f}{%
\index{ \code{-f}}%
Computes the set of reachable states before evaluating CTL
expressions.  Since NuSMV-2.4.0 this option is set by default, and it
is provided for backward compatibility only. See also option -df. }

\opt{-r}{%
\index{ \code{-r}}%
Prints the number of reachable states before exiting. If the \commandopt{f}
option is not used, the set of reachable states is computed.}

\opt{-df}{%
\index{ \code{-f}}%
Disable the computation of the set of reachable states. This option is
provided since NuSMV-2.4.0 to prevent the computation of reachable
states that are otherwise computed by default. }


\opt{-flt}{%
\index{ \code{-flt}}%
Forces the computation of the set of reachable states for the tableau
resulting from BDD-based LTL model checking (command
\command{check\_ltlspec}). If the option \commandopt{flt} is not specified 
(default), the resulting tableau will inherit the computation of the
reachable states from the model, if enabled. If the option
\commandopt{flt} is specified, the reachable states set will be calculated
for the model \emph{and} for the tableau resulting from LTL model
checking. This might improve performances of the command
\command{check\_ltlspec}, but may also lead to a dramatic slowing
down. This options has effect only when the calculation of reachable
states is enabled (see \commandopt{f}).}

\opt{-AG}{%
\index{ \code{-AG}}%
Verifies only AG formulas using an ad hoc algorithm (see documentation
for the \envvar{ag\_only\_search} environment variable).}

\opt{-coi}{%
\index{ \code{-coi}}%
%
Enables cone of influence reduction.
%
Sets to ``1'' the \envvar{cone\_of\_influence} environment variable. }
%
We remark that, when cone of influence reduction is enabled, a
counter-example trace for a property that does not hold may not be a
valid counter-example trace for the original model. We refer the
reader to the Frequently Asked Questions (FAQ)~\cite{FAQ}.

\opt{-i {\it iv\_file}}{%
\index{ \code{-i} {\it iv\_file}}%
Reads the variable ordering from file {\it iv\_file}. }

\opt{-o {\it ov\_file}}{%
\index{ \code{-o} {\it ov\_file}}%
Writes the variable ordering to file {\it ov\_file}.}

\opt{-t {\it tv\_file}}{%
\index{ \code{-t} {\it tv\_file}}%
Reads a variable list from file {\it tv\_file}. This list defines the
order for clustering the transition relation. This feature has been
provided by Wendy Johnston, University of Queensland. The results of
Johnston's et al. research have been presented at FM 2006 in Hamilton,
Canada. See \cite{fm06}.}

\opt{-reorder}{%
\index{ \code{-reorder}}%
Enables variable reordering after having checked all the specification if
any.}

\opt{-dynamic}{%
\index{ \code{-dynamic}}%
Enables dynamic reordering of variables}

\end{nusmvTable}

\begin{nusmvTable}

\opt{-m  {\it method}}{%
\index{ \code{-m} {\it method}}%
Uses {\it method} when variable ordering is enabled. Possible values for
method are those allowed for the \envvar{reorder\_method} environment
variable (see \sref{Interface to DD package}).}

\opt{-disable\_sexp2bdd\_caching}{%
\index{ \code{-disable\_sexp2bdd\_caching}}%
Sets the default value of environment variable 
\envvar{enable\_bdd\_cache} to \varvalue{0}, i.e.\@ the evaluation 
of symbolic expression to ADD and BDD representations are not cached.
See command \command{clean\_sexp2bdd\_cache} for reasons of why BDD cache
should be disabled sometimes.}

\opt{-bdd\_soh {\it heuristics}}{%
\index{ \code{-bdd\_soh}}%
Sets the default value of environment variable 
\envvar{bdd\_static\_order\_heuristics} to \varvalue{heuristics},
i.e.\@ the option sets up the heuristics to be used to compute BDD
ordering statically by analyzing the input model.  See the
documentation about variable \command{bdd\_static\_order\_heuristics}
on page \pageref{bdd_static_order_heuristics} for more details.}

\opt{-mono}{%
\index{ \code{-mono}}%
Enables monolithic transition relation}

\opt{-thresh {\it cp\_t}}{%
\index{ \code{-thresh} {\it cp\_t}}%
conjunctive partitioning with threshold of each
partition set to {\it cp\_t} (DEFAULT, with {\it cp\_t}=1000)}

\opt{-cp {\it cp\_t}}{%
\index{ \code{-cp} {\it cp\_t}}%
DEPRECATED: use \command{thresh} instead.}

\opt{-iwls95 {\it cp\_t}}{%
\index{ \code{-iwls95} {\it cp\_t}}%
Enables Iwls95 conjunctive partitioning and sets
the threshold of each partition to {\it cp\_t}}

\opt{-noaffinity}{%
\index{ \code{-noaffinity}}%
Disables affinity clustering }

\opt{-iwls95preoder}{%
\index{ \code{-iwls95preorder}}%
Enables \Iwls preordering}

\opt{-bmc}{%
\index{ \code{-bmc}}%
Enables BMC instead of BDD model checking (works only for LTL
properties and PSL properties that can be translated into LTL)}

\opt{-bmc\_length {\it k}}{%
\index{ \code{-bmc\_length} {\it k}}%
Sets \envvar{bmc\_length} variable, used by BMC}

\opt{-sat\_solver {\it name}}{%
\index{ \code{-sat\_solver} {\it name}}%
Sets \envvar{sat\_solver} variable, used by BMC so select the sat
solver to be used.}

\opt{-sin {\it on,off}}{%
\index{\code{-sin} {\it on,off}}%
Enables (on) or disables (off) Sexp inlining, by setting system
variable \varName{sexp\_inlining}. Default value is
\varvalue{off}.}

\opt{-rin {\it on,off}}{%
\index{\code{-rin} {\it on,off}}%
Enables (on) or disables (off) RBC inlining, by setting system
variable \varName{rbc\_inlining}. Default value is
\varvalue{on}. The idea about inlining was taken from
\cite{abdulla00symbolic} by Parosh Aziz Abdulla, Per Bjesse and
Niklas E\'en.}

\opt{-ojeba {\it algorithm}}{%
\index{\code{-ojeba} {\it algorithm}}%
Sets the algorthim used for BDD-based language emptiness of B\"uchi fair transition systems by setting system variable \varName{oreg\_justice\_emptiness\_bdd\_algorithm} (default is \varvalue{EL\_bwd}). The available algorithms are: \varvalue{EL\_bwd} \varvalue{EL\_fwd}}

\end{nusmvTable}
