% -*-latex-*-

In this chapter we present the syntax and semantics of the input
language of \nusmv.

Before going into the details of the language, let us give a few general
notes about the syntax.
%
In the syntax notations used below, syntactic categories
(non-terminals) are indicated by \grammar{monospace font}, and tokens
and character set members (terminals) by \textbf{bold font}.
%
Grammar productions enclosed in square brackets (`\grammar{[]}') are
optional while a vertical bar (`\grammar{|}') is used to separate
alternatives in the syntax rules. Sometimes \grammar{one of} is used
at the beginning of a rule as a shorthand for choosing among several
alternatives.
%
If the characters \grammar{\textbf{|}}, \grammar{\textbf{[}} and
\grammar{\textbf{]}} are in bold font, they lose their special
meaning and become regular tokens.

In the following, an \grammar{identifier} may be any sequence of
characters starting with a character in the set
\{{\grammar{\textbf{A-Za-z\_}}}\}
%
and followed by a possibly empty sequence of characters belonging to
the set
%
\{{\grammar{\textbf{A-Za-z0-9\_\$\#-}}}\}.
%
All characters and case in an identifier are significant. Whitespace
characters are space (\spc), tab (\tab) and newline (\ret).
%
Any string \index{comments in \nusmv language} starting with two
dashes (`\code{--}') and ending with a newline is a comment and
ignored by the parser.

\index{syntax rules!identifiers}
The syntax rule for an \grammar{identifier} is:

\begin{Grammar}
identifier :: 
        identifier_first_character
      | identifier identifier_consecutive_character

identifier_first_character :: \emph{one of}
        \textbf{A B C D E F G H I J K L M N O P Q R S T U V W X Y Z}
        \textbf{a b c d e f g h i j k l m n o p q r s t u v w x y z _}

identifier_consecutive_character :: 
        identifier_first_character
      | digit
      | \emph{one of} \textbf{\$ \# -}

digit :: \emph{one of} \textbf{0 1 2 3 4 5 6 7 8 9}
\end{Grammar}

\index{keywords}%
An \grammar{identifier} is always distinct from the \nusmv language
reserved keywords which are: \label{list of reserved keywords}\\
\begin{quote}
%
\reserved{MODULE}, \reserved{DEFINE}, \reserved{MDEFINE},
\reserved{CONSTANTS}, 
\reserved{VAR}, \reserved{IVAR}, \reserved{FROZENVAR}, \reserved{INIT},
\reserved{TRANS}, \reserved{INVAR},
\reserved{SPEC}, \reserved{CTLSPEC}, \reserved{LTLSPEC}, \reserved{PSLSPEC},
\reserved{COMPUTE}, \reserved{NAME},
\reserved{INVARSPEC}, \reserved{FAIRNESS}, \reserved{JUSTICE},
\reserved{COMPASSION},
\reserved{ISA}, \reserved{ASSIGN}, \reserved{CONSTRAINT},
\reserved{SIMPWFF}, \reserved{CTLWFF}, \reserved{LTLWFF}, \reserved{PSLWFF},
\reserved{COMPWFF}, \reserved{IN}, \reserved{MIN}, \reserved{MAX},
\reserved{MIRROR}, \reserved{PRED}, \reserved{PREDICATES},
\reserved{process}, \reserved{array}, \reserved{of}, \reserved{boolean},
\reserved{integer}, \reserved{real}, 
\reserved{word},
\operator{word1}, \operator{bool}, \operator{signed}, \operator{unsigned},
\operator{extend}, \operator{resize}, \operator{sizeof}, \operator{uwconst}, \operator{swconst},
\reserved{EX}, \reserved{AX},
\reserved{EF},
\reserved{AF}, \reserved{EG}, \reserved{AG}, \reserved{E},
\reserved{F}, \reserved{O}, \reserved{G}, \reserved{H}, \reserved{X},
\reserved{Y}, \reserved{Z}, \reserved{A}, \reserved{U}, \reserved{S},
\reserved{V}, \reserved{T}, \reserved{BU}, \reserved{EBF},
\reserved{ABF}, \reserved{EBG}, \reserved{ABG}, \reserved{case},
\reserved{esac}, \operator{mod}, \operator{next}, \reserved{init},
\operator{union}, \operator{in}, \operator{xor}, \operator{xnor},
\reserved{self}, \reserved{TRUE}, \reserved{FALSE}, \reserved{count}
\end{quote}

To represent various values we will use \grammar{integer numbers}
which are any non-empty sequence of decimal digits preceded by an
optional unary minus

\begin{Grammar}
integer_number ::
        \textbf{-} digit
      | digit
      | integer_number digit                        
\end{Grammar}

\noindent and \grammar{symbolic constants} which are \grammar{identifiers}

\index{syntax rules!symbolic constants}
\begin{Grammar}
  symbolic_constant :: identifier
\end{Grammar}

Examples of \grammar{integer numbers} and \grammar{symbolic constants}
are \code{3, -14, 007, OK, FAIL, waiting, stop}. 
%
The values of \grammar{symbolic constants} and \grammar{integer
numbers} do not intersect.


%-------------------- TYPES -----------------------------------

\section{Types Overview}
\label{Types}
\index{types}

This section provides an overview of the types that are recognised by
\nusmv.

%-----  BOOLEAN ------
\subsection{Boolean}
\label{Boolean Type}
\index{types!boolean}
\index{boolean type}

The \Boolean type comprises symbolic values \reserved{FALSE} and
\reserved{TRUE}.

%------ INTEGER -----
\subsection{Integer}
\label{Integer Type}
\index{integer type}
\index{types!integer}

% ONCE ACTUAL INTEGER TYPE IS IMPLEMENTED:
%The integer type is simply any whole number, positive or negative and
%is declared as follows:\\
%
%\indent\code{VAR a :~Integer;}\\

The domain of the \Integer type is simply any whole number, positive
or negative.
%
At the moment, there are implementation-dependent constraints on the
this type and \grammar{integer numbers} can only be in the range
$-2^{31}+1$ to $2^{31}-1$ (more accurately, these values are
equivalent to the C/C++ macros \code{INT\_MIN}+1 and \code{INT\_MAX}).

%------ ENUMERATIONS -----
\subsection{Enumeration Types}
\label{Enumeration Types}
\index{enumeration types}
\index{types!enumerations}

An \Enum type is a type specified by full enumerations of all the values
that the type comprises.  For example, the enumeration of values may be
\code{\{stopped, running, waiting, finished\}}, \code{\{2, 4, -2,
0\}}, \code{\{FAIL, 1, 3, 7, OK\}}, etc.  All elements of an enumeration have
to be unique although the order of elements is not important.

However, in the \nusmv type system, expressions cannot be of actual
\Enum types, but of their simplified and generalised versions
only. Such generalised \Enum types do not contain information about
the exact values constituting the types, but only the flag whether all
values are  \grammar{integer numbers},
\grammar{symbolic constants} or both. Below only generalised versions
of \Enum types are explained.

The \SymbEnum type covers enumerations containing only \grammar{symbolic
constants}. For example, the enumerations \code{\{stopped, running, waiting\}}
and \code{\{FAIL, OK\}} belong to the \SymbEnum type.

There is also a \IntSymbEnum type. 
This type comprises enumerations which contain \emph{both}
\grammar{integer numbers} \emph{and} \grammar{symbolic constants}, for
example, \code{\{-1, 1, waiting\}}, \code{\{0, 1, OK\}},
\code{\{running, stopped, waiting, 0\}}.

Another \Enum type is \IntEnum.
Example of enumerations of integers are \code{\{2, 4, -2, 0\}} and \code{\{-1, 1\}}. 
In the \nusmv type system an expression of the type \IntEnum
is always converted to the type \Integer. Explaining
the type of expression we will always use the type \Integer
instead of \IntEnum.

Enumerations cannot contain any boolean value (i.e.\code{\{FALSE, TRUE\}}).
\Boolean type must be declared as boolean.

To summarise, we actually deal only with two \Enum types: \SymbEnum and
\IntSymbEnum. These types are distinguishable and have different operations
allowed on them.


%------ WORD  -----
\subsection{Word}
\label{Word Type}
\index{word type}
\index{types!word}

The \UWord and \SWord types are used to model vector of bits (booleans) which
allow bitwise logical and arithmetic operations (unsigned and signed, respectively).
%
These types are distinguishable by their width. For example, type
\UWord[3] represents vector of three bits, which allows unsigned operations, and type
\SWord[7] represents vector of seven bits, which allows signed operations.

When values of \UWord[N] are interpreted as integer numbers the bit
representation used is the most popular one, i.e.\@ each bit represents
a successive power of 2 between $0$ (bit number 0) and $2^{N-1}$ (bit
number $N-1$). Thus \UWord[N] is able to represent values from 0 to
$2^{N}-1$. 

The bit representation of \SWord[N] type is ``two's complement'',
i.e.\@ it is the same as for \UWord[N] except that the highest bit
(number $N-1$) has value $-2^{N-1}$. Thus the possible value for
\SWord[N] are from $-2^{N-1}$ to $2^{N-1}-1$.


%------ ARRAY -----
\subsection{Array}
\label{Array Type}
\index{array type}
\index{types!array}

Arrays are declared with a lower and upper bound for the index, and
the type of the elements in the array. For example,\\

\indent\code{\Array 0..3 of \Boolean}\\
\indent\code{\Array 10..20 of \{OK, y, z\}}\\
\indent\code{\Array 1..8 of \Array~-1..2 of \UWord[5]}\\

The type \code{\Array 1..8 of \Array~-1..2 of \UWord[5]} means an
array of 8 elements (from 1 to 8), each of which is an array of 4
elements (from -1 to 2) that are 5-bit-long unsigned words.

Array subtype is the immediate subtype of an array type. For example,
subtype of \code{\Array 1..8 of \Array~-1..2 of \UWord[5]}
is \code{\Array~-1..2 of \UWord[5]} which has its own subtype 
\UWord[5].

\Array types are incompatible with \Set type, i.e.\@ array elements cannot 
be of \Set type.

Expression of array type can be constructed with array
\reserved{DEFINE} (see \ref{Array Define Declarations}) or variables of
array type (see \ref{Type Specifiers}).

% For information about arrays of modules, see \sref{???}.

% %------ WORD ARRAY -----
% \subsection{WordArray}
% \label{WordArray Type}
% \index{word array type}
% \index{types!word array}
% 
% The \WordArray types are used to model arrays whose size and
% elements are specified with \Word types.
% For example, \\
% \indent\code{array word[5] of unsigned word[3];}\\
% \indent\code{array word[4] of unsigned word[9];}
% 
% The type \code{array word[4] of word[9]}, i.e. \WordArray[{[4][9]}],
% means an array of 16 elements (from 0d4\_0 to 0b4\_15), each of which
% is \UWord[9].
% \WordArray types are distinguishable on their size and width of elements.
% Note also that both the size and width of elements have to be greater
% than zero. 
% 
% \WordArray are very specific typse and very few operators can 
% be applied to expressions of these types. See ...  only WAREAD, WAWRITE,
% := , = operators.
 
%------ SET-TYPES -----
\subsection{Set Types}
\label{Set Types}
\index{set types}
\index{types!set}

\Set types are used to identify expressions representing a set of values.
There are four \Set types: \BoolSet, \IntSet, \SymbSet, \IntSymbSet.
The \Set types can be used in a very limited number of ways. In
particular, a variable cannot be of a \Set type. Only \grammar{range
constant} and \operator{union} operator can be used to create an
expression of a \Set type, and only
\operator{in}, \operator{case}, \itebullet and assignment\footnote{For more information
on these operators see pages \pageref{Range Constant},
\pageref{Set Expressions},
\pageref{Inclusion Operator}, \pageref{Case Expressions} and 
\pageref{ASSIGN Constraint}, respectively.} expressions can have imediate operands of a \Set
type.

Every \Set type has a counterpart among other types. In particular,
\begin{itemize}
\item[] the counterpart of a \BoolSet type is \Boolean, 
\item[] the counterpart of a \IntSet type is \Integer, 
\item[] the counterpart of a \SymbSet type is \SymbEnum, 
\item[] the counterpart of a \IntSymbSet type is \IntSymbEnum. 
\end{itemize}
Some types such as \UWord and \SWord %and \WordArray 
do not have a \Set type counterpart.


%------ TYPE ORDER  -----
\subsection{Type Order}
\label{Type Order}
\index{type order}
\index{types!ordering}
%
Figure~\ref{fig:typehierarchy} depicts the order existing between
types in \nusmv.

\begin{figure}[h]
\begin{center}
\begin{tabular}{ccl} %@{\qquad}
% boolean type
      \begin{tabular}{cc}
        \Boolean\\
      \end{tabular}
% scalar types
      \begin{tabular}{cc}
        \Integer & \SymbEnum\\
        $\downarrow$ & $\downarrow$\\
        \multicolumn{2}{c}{\IntSymbEnum}\\
      \end{tabular} & &
% set types
      \begin{tabular}{c}
        \UWord[1]\\
        \\
        \raisebox{1.3ex}[0pt]{~\UWord[2]}\\
        \UWord[3]\\
        \ldots \\
      \end{tabular} \\\\\\
% unsigned words
      \begin{tabular}{cc}
        \BoolSet\\
      \end{tabular}
      \begin{tabular}{cc}
        \IntSet & \SymbSet\\
        $\downarrow$ & $\downarrow$\\
        \multicolumn{2}{c}{\IntSymbSet}\\
      \end{tabular} & &
% signed words
      \begin{tabular}{c}
        \SWord[1]\\
        \\
        \raisebox{1.3ex}[0pt]{~\SWord[2]}\\
        \SWord[3]\\
        \ldots \\
      \end{tabular} \\\\\\
% array types 
      \begin{tabular}{c}
        \Array N1..M1 of subtype1\\
        $\downarrow$\\
        \Array N2..M2 of subtype2\\
      \end{tabular} & 
      \begin{tabular}{c}
       \\if and only if\\
      \end{tabular} & 
      \begin{tabular}{cc}
        N1=N2      & subtype1\\
        M1=M2, and & $\downarrow$\\
                   & subtype2\\
      \end{tabular} 
%
%      \begin{tabular}{c}
%        \WordArray[{[1][1]}]\\
%        \\
%        \raisebox{1.3ex}[0pt]{~\WordArray[{[1][2]}]}\\
%        \WordArray[{[1][3]}]\\
%        \ldots \\
%      \end{tabular}
\end{tabular}
\end{center}
\caption{The ordering on the types in \nusmv\label{fig:typehierarchy}}
\end{figure}

\noindent It means, for example, that \Integer is less 
than \IntSymbEnum, \SymbEnum is less than \IntSymbEnum, etc. 
%
The \UWord and \SWord %and \WordArray types are not comparable with
any other type or between each other. Any type is equal to itself.

Note that enumerations containing only \grammar{integer numbers} have
the type \Integer.


For 2 arrays types \code{array N1..M1 of subtype1} and \code{array
N2..M2 of subtype2} the first type is less then the second one if and
only if \code{N1=N2}, \code{M1=M2} and type \code{subtype1} is less
than \code{subtype2}.

%-------------------- EXPRESSIONS -----------------------------------

\section{Expressions}
\label{Expressions}
\index{expressions}
%

The previous versions of NuSMV (prior to 2.4.0) did not have the type
system and as such expressions were untyped. 
%
In the current version all expressions are typed and there are
constraints on the type of operands. 
%
Therefore, an expression may now potentially violate the type system,
i.e. be erroneous.

To maintain backward compatibility, there is a new system variable
called \code{backward\_compatibility} (and a correponding \code{-old}
command line option) that disables a few new features of version 2.4
to keep backward compatibility with old version of
\nusmv. In particular, if this system variable is set then type violations 
caused by expressions of old types (i.e. \Enum type, \Boolean and
\Integer) will be ignored by the type checker, instead, warnings will
be printed out. See description at page
\pageref{ref::backwardcompatibility} for further information.

If additionally, the system variable
\code{type\_checking\_warning\_on} is
\emph{un}set, then even these warnings will not be printed out.

%-----  IMPLICIT TYPE CONVERSION  ------
\subsection{Implicit Type Conversion}
\label{Implicit Type Conversion}
\index{implicit type conversion}
\index{types!implicit conversion}

In some expressions operands may be converted from one type to
its \Set type counterpart (see \ref{Set Types}). For example, \Integer
can be converted to \IntSet type.

\textbf{\textit{Note:}} Prior to version 2.5.1, implicit type 
conversion from \Integer to \Boolean (and viceversa) was
performed. Since version 2.5.1, implicit \Integer <-> \Boolean type
conversion is no longer supported, and explicit cast operators have
to be used.

%-----  CONSTANT EXPRESSION  ------
\subsection{Constant Expressions}
\label{Constant Expressions}
\index{constant expressions}
\index{expressions!constants}%
%
A \grammar{constant} can be a boolean,
integer, symbolic, word or range constant.

\begin{Grammar}
constant ::
        boolean_constant
      | integer_constant
      | symbolic_constant
      | word_constant
      | range_constant
\end{Grammar}

%----- BOOLEAN CONSTANT   ------
\subsubsection{Boolean Constant}
\label{Boolean Constant}

A \grammar{boolean constant} is one of the symbolic
values \reserved{FALSE} and \reserved{TRUE}.
%
The type of a \grammar{boolean constant} is \Boolean.

\begin{Grammar}
boolean_constant :: \emph{one of}
        \reserved{FALSE} \reserved{TRUE}
\end{Grammar}

%----- INTEGER CONSTANT   ------
\subsubsection{Integer Constant}
\label{Integer Constant}

An \grammar{integer constant} is an \grammar{integer number}. The type
of an \grammar{integer constant} is \Integer.

\begin{Grammar}
integer_constant :: integer_number
\end{Grammar}

%----- SYMBOLIC CONSTANT   ------
\subsubsection{Symbolic Constant}
\label{Symbolic Constant}

A \grammar{symbolic constant} is syntactically an \grammar{identifier}
and indicates a unique value.

\begin{Grammar}
symbolic_constant :: identifier
\end{Grammar}

\noindent The type of a \grammar{symbolic constant} is \SymbEnum. 
%
See \sref{Namespaces} for more information about how
\grammar{symbolic constants} are distinguished from other
\grammar{identifiers}, i.e. variables, defines, etc.

%----- WORD CONSTANT  ------
\subsubsection{Word Constant}
\label{Word Constant}

\grammar{Word constant} begins with digit \code{0}, 
followed by optional character \code{u} (unsigned) or \code{s}
(signed) and one of the characters \code{b}/\code{B} (binary),
\code{o}/\code{O} (octal), \code{d}/\code{D} (decimal) or 
\code{h}/\code{H} (hexadecimal) which 
gives the base that the actual constant is in. 
%
Next comes an optional decimal integer giving the number of bits, 
then the character \code{\_}, and lastly the constant value itself.
%
Assuming \code{N} is the width of the constant the type of a
\grammar{word constant} is \SWord[N] if character \code{s} is
provided, and \UWord[N] otherwise.  For example:

\begin{center}
\begin{tabular}{l@{\  has type\  }l}
\code{0sb5\_10111} & \SWord[5]\\
\code{0uo6\_37} & \UWord[6]\\
\code{0d11\_9} & \UWord[11]\\
\code{0sh12\_a9} & \SWord[12]\\
\end{tabular}
\end{center}

\noindent The number of bits can be skipped, in which case the
width is automatically calculated from the number of digits in the
constant and its base. 
%
It may be necessary to explicitly give leading zeroes to make the type
correct --- the following are all equivalent declarations of the
integer constant \code{11} as a word of type \UWord[8]:

\begin{center}
\begin{tabular}{l}
    \code{0ud8\_11}\\
    \code{0ub8\_1011}\\
    \code{0b\_00001011}\\
    \code{0h\_0b}\\
    \code{0h8\_b}\\
\end{tabular}
\end{center}

\noindent The syntactic rule of the \grammar{word constant} is the
following:

\begin{Grammar}
word_constant :: 
        \textbf{0} [word_sign_specifier] word_base [word_width] \textbf{_} word_value

word_sign_specifier :: \emph{one of} 
        \textbf{u s}

word_width :: 
        integer_number       -- a number greater than zero

word_base :: 
        \textbf{b} | \textbf{B} | \textbf{o} | \textbf{O} | \textbf{d} | \textbf{D} | \textbf{h} |  \textbf{H}

word_value :: 
        hex_digit
      | word_value hex_digit
      | word_value \textbf{\_}

hex_digit :: \emph{one of}  
        \textbf{0 1 2 3 4 5 6 7 8 9 a b c d e f A B C D E F}
\end{Grammar}

\label{the notes on word constants}
\noindent Note that 

\begin{itemize}
  \item The width of a word must be a number strictly greater than 0. 
  \item Decimal \grammar{word constants} \textit{must} be declared
        with the width specifier, since the number of bits needed for
        an expression like \code{0d\_019} is unclear.
  \item Digits are restricted depending on the base the constant is
        given in.
  \item Digits can be separated by the underscore character
        ("{\code{\_}}") to aid clarity, for example
        \code{0b\_0101\_1111\_1100} which is equivalent to
        \code{0b\_010111111100}.

 \item For a given width \code{N} the value of a constant has to be in
       range $0 \ldots 2^{N}-1$. For decimal signed words (both \code{s}
       and \code{d} are provided) the value of a constant has to be in
       range $0 \ldots 2^{N-1}$.

 \item The number of bits in \grammar{word constant} has an implementation
       limit which for most systems is 64 bits.
\end{itemize}

%-----  RANGE CONSTANT  ------
\subsubsection{Range Constant}
\label{Range Constant}

A \grammar{range constant} specifies a set of consecutive integer
numbers. For example, a constant \code{-1..5} indicates the set of
numbers \code{ -1, 0, 1, 2, 3, 4} and \code{5}. Other examples of
\grammar{range constant} can be \code{1..10},\ \ \code{-10..-10},
\ \ \code{1..300}. 
The syntactic rule of the \grammar{range constant} is the following:
\begin{Grammar}
range_constant :: 
        integer_number \textbf{..} integer_number
\end{Grammar}
with an additional constraint that the first integer number must be
less than or equal to the second integer number.
%
The type of a \grammar{range constant} is \IntSet.

%-----  BASIC EXPRESSION   ------
\subsection{Basic Expressions}
\label{Basic Expressions}
\index{expressions!basic expressions}


A basic expression is the most common kind of expression used in
\nusmv.

\begin{Grammar}
basic_expr ::
      constant                      -- a constant
    | variable_identifier           -- a variable identifier
    | define_identifier             -- a define identifier
    | \textbf{(} basic_expr \textbf{)}
    | \operator{!} basic_expr                  -- logical or bitwise NOT
    | basic_expr \operator{\&} basic_expr       -- logical or bitwise AND
    | basic_expr \operator{|} basic_expr       -- logical or bitwise OR
    | basic_expr \reserved{xor} basic_expr     -- logical or bitwise exclusive OR
    | basic_expr \reserved{xnor} basic_expr    -- logical or bitwise NOT exclusive OR
    | basic_expr \operator{->} basic_expr      -- logical or bitwise implication
    | basic_expr \operator{<->} basic_expr     -- logical or bitwise equivalence
    | basic_expr \operator{=} basic_expr       -- equality
    | basic_expr \operator{!=} basic_expr      -- inequality
    | basic_expr \operator{<} basic_expr       -- less than
    | basic_expr \operator{>} basic_expr       -- greater than
    | basic_expr \operator{<=} basic_expr      -- less than or equal
    | basic_expr \operator{>=} basic_expr      -- greater than or equal
    | \operator{-} basic_expr                  -- integer unary minus
    | basic_expr \operator{+} basic_expr       -- integer addition
    | basic_expr \operator{-} basic_expr       -- integer subtraction
    | basic_expr \operator{*} basic_expr       -- integer multiplication
    | basic_expr \operator{/} basic_expr       -- integer division
    | basic_expr \reserved{mod} basic_expr     -- integer remainder
    | basic_expr \operator{>>} basic_expr      -- bit shift right
    | basic_expr \operator{<<} basic_expr      -- bit shift left
    | basic_expr \textbf{[} index \textbf{]}          -- index subscript
    | basic_expr \textbf{[} basic_expr \textbf{:} basic_expr \textbf{]}
                                    -- word bits selection
    | basic_expr \operator{::} basic_expr      -- word concatenation
    | \operator{word1} \textbf{(} basic_expr \textbf{)}          -- boolean to unsigned word[1] conversion
    | \operator{bool} \textbf{(} basic_expr \textbf{)}           -- unsigned word[1] and int to boolean conversion
    | \operator{toint} \textbf{(} basic_expr \textbf{)}          -- word and boolean to integer constant conversion
    | \reserved{count} \textbf{(} basic_expr_list \textbf{)}     -- count of true boolean expressions
    | \operator{swconst} \textbf{(} basic_expr , basic_expr \textbf{)}        
                                    -- integer to signed word constant conversion
    | \operator{uwconst} \textbf{(} basic_expr, basic_expr \textbf{)}        
                                    -- integer to unsigned word constant conversion
    | \operator{signed} \textbf{(} basic_expr \textbf{)}         -- unsigned word to signed word conversion
    | \operator{unsigned} \textbf{(} basic_expr \textbf{)}       -- signed word to unsigned word conversion
    | \operator{sizeof} \textbf{(} basic_expr \textbf{)}         -- word size as an integer
    | \operator{extend} \textbf{(} basic_expr , basic_expr\textbf{)}  
                                    -- word width extension
    | \operator{resize} \textbf{(} basic_expr , basic_expr\textbf{)}  
                                    -- word width resize
    | basic_expr \operator{union} basic_expr   -- union of set expressions 
    | \textbf{\{} set_body_expr \textbf{\}}             -- set expression
    | basic_expr \operator{in} basic_expr      -- inclusion in a set expression
    | basic_expr \operator{?} basic_expr \operator{:} basic_expr      
                                    -- if-then-else expression 
    | case_expr                     -- case expression
    | basic_next_expr               -- next expression

basic_expr_list ::
      basic_expr
    | basic_expr_list \textbf{,} basic_expr
\end{Grammar}
%    | basic_expr \operator{>>>} basic_expr     -- bit rotation right
%    | basic_expr \operator{<<<} basic_expr     -- bit rotation left

\index{operators!precedence}
\noindent The order of parsing precedence for operators from high to
low is:
%
\begin{Grammar}
      \operator{[ ]}, \operator{[ : ]}
      \operator{!}
      \operator{::}
      \operator{-} (unary minus)
      \operator{*}   \operator{/}   \operator{mod}
      \operator{+}   \operator{-}
      \operator{<<}   \operator{>>}
      \reserved{union}
      \reserved{in}
      \operator{=}   \operator{!=}   \operator{<}   \operator{>}   \operator{<=}   \operator{>=}
      \operator{\&}
      \operator{|}   \operator{xor}   \operator{xnor}
      \itebullet
      \operator{<->}
      \operator{->}
\end{Grammar}
%\operator{<<<}   \operator{>>>}
%
Operators of equal precedence associate to the left, except \operator{->}
that associates to the right.
%
The constants and their types are explained in \sref{Constant Expressions}.

%----- VARIABLES AND DEFINES   ------
\subsubsection{Variables and Defines}
\label{Variables and Defines}
\index{defines}
\index{variables}

A \grammar{variable\_identifier} and \grammar{define\_identifier} are
expressions which identify a variable or a define, respectively. 
%
Their syntax rules are:

\begin{Grammar}
define_identifier :: complex_identifier

variable_identifier :: complex_identifier
\end{Grammar}
%
The syntax and semantics of \grammar{complex\_identifiers} are explained
in \sref{References to Module Components}.  
%
All defines and variables referenced in expressions should be
declared. All identifiers (variables, defines, symbolic constants,
etc) can be used prior to their definition, i.e. there is no constraint on
order such as in C where a declaration of a variable should always be
placed in text above the variable use.
%
See more information about define and variable declarations in
\sref{DEFINE Declarations} and \sref{Variable Declarations}.

A define is a kind of macro. 
%
Every time a define is met in expressions, it is substituted by the
expression associated with this define. 
%
Therefore, the type of a define is the type of the associated
expression in the current context.
%
Define expressions may contain \operator{next} operators; Normal rules
apply: No nested \operator{next} operators.

\grammar{variable\_identifier} represents state, input, and frozen
variables. 
%
The type of a variable is specified in its declaration. 
%
For more information about variables, see \sref{Definition of the
FSM}, \sref{State Variables}, \sref{Input Variables}, and
\sref{Frozen Variables}.
%
Since a \grammar{symbolic constant} is syntactically indistinguishable
from \grammar{variable\_iden\-tifiers} and
\grammar{define\_identifiers}, a symbol table is used to distinguish
them from each other.

%-----  PARENTHESES  ------
\subsubsection{Parentheses}
\label{Parentheses}
\index{parentheses}

Parentheses may be used to group expressions. 
%
The type of the whole expression is the same as the type of the
expression in the parentheses.

%-----    ------
\subsubsection{Logical and Bitwise !}
\label{Logical and Bitwise NOT}
\index{NOT! logical and bitwise}
\index{operators!NOT}

The \emph{signature} of the logical and bitwise NOT operator
\operator{!} is:\\

\begin{tabular}{l@{ : }l}
\operator{!} & \Boolean $\rightarrow$ \Boolean\\
             & \UWord[N] $\rightarrow$ \UWord[N]\\
             & \SWord[N] $\rightarrow$ \SWord[N]\\
\end{tabular}\\

\noindent This means that the operation can be applied to \Boolean,
\UWord and \SWord operands. 
%
The type of the whole expression is the same as the type of the
operand.
%
If the operand is not \Boolean, \UWord or \SWord then the expression
violates the type system and \nusmv will throw an error.


%-----    ------
\subsubsection{Logical and Bitwise \operator{\&}, \operator{|}, \operator{xor}, \operator{xnor}, \operator{->}, \operator{<->}}
\label{Logical and Bitwise AND, OR, XOR, XNOR, IMPLIES and IFF}
\index{AND! logical and bitwise}
\index{OR! logical and bitwise}
\index{XOR! logical and bitwise}
\index{XNOR! logical and bitwise}
\index{IMPLIES! logical and bitwise}
\index{IFF! logical and bitwise}
\index{operators!AND}
\index{operators!OR}
\index{operators!XOR}
\index{operators!XNOR}
\index{operators!IMPLIES}
\index{operators!IFF}

Logical and bitwise binary operators \operator{\&} (AND), \operator{|}
(OR), \operator{xor} (exclusive OR), \operator{xnor} (negated
exclusive OR), \operator{->} (implies) and \operator{<->} (if and only
if) are similar to the unary operator \operator{!}, except that they
take two operands.
%
Their signature is:\\

\noindent
\begin{tabular}{l@{ : }l}
\operator{\&}, \operator{|}, \operator{xor}, \operator{xnor}, \operator{->}, \operator{<->}
     & \Boolean * \Boolean $\rightarrow$ \Boolean\\
     & \UWord[N] * \UWord[N] $\rightarrow$ \UWord[N]\\
     & \SWord[N] * \SWord[N] $\rightarrow$ \SWord[N]\\
\end{tabular}\\

\noindent the operands can be of \Boolean, \UWord or \SWord type, and the type
of the whole expression is the type of the operands.  
%
Note that both word operands should have the same width.

%-----    ------
\subsubsection{Equality (\operator{=}) and Inequality (\operator{!=})}
\label{Equality and Inequality}
\index{operators!equality}
\index{operators!inequality}

The operators \operator{=} (equality) and \operator{!=} (inequality)
have the following signature:\\

\noindent
\begin{tabular}{ll}
\operator{=}, \operator{!=}
     &{ : }\Boolean * \Boolean $\rightarrow$ \Boolean\\
     &{ : }\Integer * \Integer $\rightarrow$ \Boolean\\
     &{ : }\SymbEnum  * \SymbEnum $\rightarrow$  \Boolean\\
     &{ : }\IntSymbEnum * \IntSymbEnum $\rightarrow$ \Boolean\\
     &{ : }\UWord[N] * \UWord[N] $\rightarrow$ \Boolean\\
     &{ : }\SWord[N] * \SWord[N] $\rightarrow$ \Boolean\\
%     &{ : }\WordArray[{[N][M]}] * \WordArray[{[N][M]}] $\rightarrow$ \Boolean\\
\end{tabular}\\

No implicit type conversion is performed. 
%
For example, in the expression\\
%
\code{TRUE = 5}\\
%
\noindent the left operand is of type \Boolean and the right one is of
type \Integer. 
%
Though the signature of the operation does not have a \Boolean *
\Integer rule, the expression is not correct, because no implicit type
conversion will be performed. One can use the \operator{toint} or
the \operator{bool} for explicit casts. \\
%
For example:

\code{toint(TRUE) = 5}

or

\code{TRUE = bool(5)}

\noindent This is also true if one of the operands is of type \UWord[1]
and the other one is of the type \Boolean. Explicit cast must be used
(e.g. using \operator{word1} or \operator{bool})


%-----    ------
\subsubsection{Relational Operators \operator{>}, \operator{<}, \operator{>=}, \operator{<=}}
\label{Relational Operators}
\index{\operator{>},\operator{<},\operator{>=},\operator{<=}}
\index{operators!relational}

The relational operators \operator{>} (greater than), \operator{<}
(less than), \operator{>=} (greater than or equal to) and
\operator{<=} (less than or equal to) have the following signature:\\

\begin{tabular}{l@{ : }l}
\operator{>}, \operator{<}, \operator{>=}, \operator{<=}
& \Integer * \Integer $\rightarrow$ \Boolean\\
& \UWord[N] * \UWord[N] $\rightarrow$ \Boolean\\
& \SWord[N] * \SWord[N] $\rightarrow$ \Boolean\\
\end{tabular}


%-----    ------
\subsubsection{Arithmetic Operators \operator{+}, \operator{-}, \operator{*}, \operator{/}}
\label{Arithmetic Operators}
\index{operators!arithmetic}
\index{\operator{+},\operator{-},\operator{*},\operator{/}}

The arithmetic operators \operator{+} (addition), \operator{-} (unary
negation or binary subtraction), \operator{*} (multiplication) and
\operator{/} (division) have the following signature:\\

\begin{tabular}{l@{ : }l}
\operator{+}, \operator{-}, \operator{*}, \operator{/}
     & \Integer * \Integer $\rightarrow$ \Integer\\
     & \UWord[N] * \UWord[N] $\rightarrow$ \UWord[N]\\
     & \SWord[N] * \SWord[N] $\rightarrow$ \SWord[N]\vspace{10pt}\\
\operator{-}(unary) 
     & \Integer $\rightarrow$ \Integer\\
     & \UWord[N] $\rightarrow$ \UWord[N]\\
     & \SWord[N] $\rightarrow$ \SWord[N]\\

\end{tabular}\\

\noindent Before checking the expression for being correctly typed,
the implicit type conversion can be applied to \emph{one} of the
operands.
%
If the operators are applied to \UWord[N] or \SWord[N]type, then the operations
are performed modulo $2^N$.

The result of the \operator{/} operator is the quotient from the
division of the first operand by the second.
%
The result of the \operator{/} operator is
the algebraic quotient with any fractional part discarded (this is
often called ``truncation towards zero'').
%
If the quotient \code{a/b} is representable, the expression
\code{(a/b)*b + (a mod b)} shall equal \code{a}.
%
If the value of the second operand is zero, the behavior is undefined
and an error is thrown by \nusmv.
%
The semantics is equivalent to the corresponding one of C/C++
languages.

In the versions of \nusmv prior to 2.4.0 the semantics of division was
different. See page \pageref{Old division semantics} for more detail.


%-----    ------
\subsubsection{Remainder Operator \operator{mod}}
\label{Remainder Operator}
\index{operator!mod}
\index{\operator{mod}}

The result of the \operator{mod} operator is the
algebraic remainder of the division.
%
If the value of the second operand is zero, the behavior is undefined
and an error is thrown by \nusmv.

The signature of the remainder operator is:\\

\begin{tabular}{l@{ : }l}
\operator{mod} & \Integer * \Integer $\rightarrow$ \Integer\\
               & \UWord[N] * \UWord[N] $\rightarrow$ \UWord[N]\\
               & \SWord[N] * \SWord[N] $\rightarrow$ \SWord[N]\\
\end{tabular}\\

\noindent The semantics of \operator{mod} operator is
equivalent to the corresponding operator \operator{\%} of C/C++
languages. Thus if the quotient \code{a/b} is representable, the
expression \code{(a/b)*b + (a mod b)} shall equal \code{a}.


\bigskip

\textbf{\textit{Note:}}\label{Old division semantics}
in older versions of \nusmv (priori 2.4.0)
the semantics of quotient and remainder were different.  Having the
division and remainder operators $/$ and $mod$ be of the current, i.e.
C/C++'s, semantics the older semantics of division was given by
the formula:\\
\indent IF (a $mod$  b $<$ 0) THEN (a $/$ b $-$ 1) ELSE (a $/$ b)\\
and the semantics of remainder operator was given by the formula:\\
\indent IF (a $mod$  b $<$ 0) THEN (a $mod$ b $+$ b) ELSE (a $mod$ b)\\
Note that in both versions the equation \code{(a/b)*b + (a mod b) = a}
holds. For example, in the current version of NuSMV the following
holds:\\
\begin{tabular}{ll}
 7/5 = 1   &  7 mod 5 = 2\\
 -7/5 = -1 & -7 mod 5 = -2\\
 7/-5=-1   &  7 mod -5 = 2\\
 -7/-5=1   & -7 mod -5 = -2
\end{tabular}\\
whereas in the older versions on NuSMV the equations were\\
\begin{tabular}{ll}
 7/5 = 1   &  7 mod 5 = 2\\
 -7/5 = -2 & -7 mod 5 = 3\\
 7/-5=-1   &  7 mod -5 = 2\\
 -7/-5=0   & -7 mod -5 = -7
\end{tabular}\\
When supplied, the command line option -old\_div\_op switches the
semantics of division and remainder to the old one.

\bigskip
\textbf{\textit{Note:}}\label{Remainder vs Modulo operator}
semantics of \emph{modulo} operator can be obtained from the
remainder operator by exploiting the equivalence:

\indent (n modulo M) $\equiv$ ((n $mod$ M) $+$ M) $mod$ M

Where $mod$ is the remainder operator.

%-----    ------
\subsubsection{Shift Operators \operator{<<}, \operator{>>}}
%\subsubsection{Shift and Rotate Operators \operator{<<}, \operator{>>}, \operator{<<<}, \operator{>>>}}
\label{Shift Operators}
\index{Shift Operator}
%\index{Rotate Operator}
%\index{operators!rotate}
\index{operators!shift}
\index{\operator{<<}, \operator{>>}}

The signature of the shift %and rotate
operators is:\\

\begin{tabular}{l@{ : }l}
\operator{<<}, \operator{>>}%, \operator{<<<}, \operator{>>>}
& \UWord[N] * \Integer $\rightarrow$ \UWord[N]\\
& \SWord[N] * \Integer $\rightarrow$ \SWord[N]\\
& \UWord[N] * \UWord[M] $\rightarrow$ \UWord[N]\\
& \SWord[N] * \UWord[M] $\rightarrow$ \SWord[N]\\
\end{tabular}\\
Before checking the expression for being correctly typed, the right
operand can be implicitly converted from \Boolean to \Integer type.

Left shift \operator{<<} (right shift \operator{>>}) operation shifts
to the left (right) the bits of the left operand by the number
specified in the right operand.
%
A shift by N bits is equivalent to N shifts by 1 bit.
%
A bit shifted behind the word bound is lost. 
%
During shifting a word is padded with zeros with the exception of
the right shift for \SWord, in which case a word is padded
with its highest bit.
%
%Rotates (\operator{<<<} and \operator{>>>}) are similar to shifts
%except that a bit shifted behind the word bound appears on the other
%side of the word. 
%
For instance,

\begin{center}
\begin{tabular}{l@{\qquad\qquad\qquad\qquad}l}
0ub4\_0101 \operator{<<} 2 is equal to & 0sb3\_1011 \operator{>>} 2 is equal to \\
0ub4\_0100 \operator{<<} 1 is equal to & 0sb3\_1110 \operator{>>} 1 is equal to \\
0ub4\_1000 \operator{<<} 0 is equal to & 0sb3\_1111 \operator{>>} 0 is equal to \\
0ub4\_1000 and                         & 0sb3\_1111\\
\end{tabular}
\end{center}


It has to be remarked that the shifting %and rotation
requires the right operand to be greater or equal to zero and less
then or equal to the width of the word it is applied to. 
%
\nusmv raises an error if a shift is attempted that does %/rotation 
 not satisfy this restriction.

%-----    ------
\subsubsection{Index Subscript Operator \operator{[~]}}
\label{Index Subscript Operator}
\index{index subscript operator}
\index{operators!index subscript}
\index{\operator{[]}}

The index subscript operator extracts one element of an array in the
typical fashion.
%
On the left of \operator{[]} there has to be an expression
of array type.
%
The index expression in the brackets has to be an expression of
\Integer or \AnyWord type with value greater or equal to lower bound
and less or equal to the upper bound of the array.
%
The signature of the index subscript operator is:\\

\noindent
\begin{tabular}{l@{ : }l}
\operator{[~]} 
& \Array N..M of subtype * \AnyWord[N] $\rightarrow$ subtype\\
& \Array N..M of subtype * \Integer $\rightarrow$ subtype \\
\end{tabular}\\

For example, for below declarations
%
\footnote{See \ref{Array Define Declarations}) for array defines and 
\ref{Type Specifiers} for array variables.}
%
:
\begin{nusmvCode}
MODULE main
 VAR a : array -1 .. 4 of array 1 .. 2 of boolean;
 DEFINE d := [[12, 4], [-1,2]];
 VAR r : 0..1;
\end{nusmvCode}
expressions \code{a[-1]}, \code{a[0][r+1]} and \code{d[r][1]} are
valid whereas \code{a[0]},  \code{a[0][r]} and \code{d[0][r-1]} will
cause out of bound error.

%-----    ------
\subsubsection{Bit Selection Operator \operator{[ :~]}}
\label{Bit Selection Operator}
\index{bit selection operator}
\index{operators!bit selection}
\index{\operator{[:]}}

The bit selection operator extracts consecutive bits from a \UWord or
\SWord expression, resulting in a new \UWord expression.
%
This operation always decreases the width of a word or leaves it
intact. 
%
The expressions in the brackets have to be integer constants which
specify the high and low bound.
%
The high bound must be greater than or equal to the low bound.
%
The bits count from 0. 
%
The result of the operations is \UWord value consisting of the
consecutive bits beginning from the high bound of the operand down to,
and including, the low bound bit.
%
For example, 0sb7\_1011001[4:1] extracts bits 1 through 4 (including 1st
and 4th bits) and is equal to 0ub4\_1100.
%
0ub3\_101[0:0] extracts bit number 0 and is equal to 0ub1\_1.\\

The signature of the bit selection operator is:\\

\noindent
\begin{tabular}{l@{ : }l}
\operator{[ :~]} 
& \UWord[N] * $\Integer_{h}$ * $\Integer_{l}$ $\rightarrow$
  \UWord[$\Integer_{h} - \Integer_{l} + 1$]\\
& \SWord[N] * $\Integer_{h}$ * $\Integer_{l}$ $\rightarrow$
  \UWord[$\Integer_{h} - \Integer_{l} + 1$]\\
\end{tabular}\\

\textit{where} $0 \leq \Integer_{l} \leq \Integer_{h} <$ \code{N}


%-----    ------
\subsubsection{Word Concatenation Operator \code{::}}
\label{Word Concatenation Operator}
\index{operators!word concatenation}
\index{concatenation operator}
\index{\operator{::}}

The concatenation operator joins two words (\UWord or \SWord or both)
together to create a larger \UWord type.
%
The operator itself is two colons (\operator{::}), and its signature
is as follows:\\

\begin{tabular}{l@{ : }l}
\operator{::} & \AnyWord[M] * \AnyWord[N] $\rightarrow$ \UWord[M+N]\\
\end{tabular}\\

\noindent where \AnyWord[N] is \UWord[N] or \SWord[N].
The left-hand operand will make up the upper bits of the new
word, and the right-hand operand will make up the lower bits. The
result is always \UWord.
%
For example, given the two words \code{w1 := 0ub4\_1101} and \code{w2
:= 0sb2\_00}, the result of \code{w1{\operator{::}}w2} is
\code{0ub6\_110100}.


%-----    ------
\subsubsection{Word sizeof Operator}
\label{Word sizeof Operator}
\index{sizeof operator}

\operator{sizeof} operator provides a very simple way for retrieving the width of a word. The behavior of this operator can be
described as follows:\\
\\
let w be a \AnyWord[N] : \operator{sizeof}(w) returns N.\\
\\
The signature of the operator is:\\

\begin{tabular}{l@{ : }l}
\operator{sizeof} 
& \UWord $\rightarrow$ \Integer\\
& \SWord $\rightarrow$ \Integer\\
\end{tabular}\\
%
%-----    ------
\subsubsection{Extend Word Conversions}
\label{Extend Word Conversions}
\index{extend operator}

\operator{extend} operator increases the width of a word by attaching 
additional bits on the left. If the provided word is unsigned then
zeros are added, otherwise if the word is signed the highest (sing)
bit is repeated corresponding number of times.

The signature of the operator is:\\

\begin{tabular}{l@{ : }l}
\operator{extend} 
& \UWord[N] * \Integer $\rightarrow$ \UWord[N+\Integer]\\
& \SWord[N] * \Integer $\rightarrow$ \SWord[N+\Integer]\\
\end{tabular}\\

\noindent For example:
%
\begin{center}
\begin{tabular}{l}
{\operator{extend}}(0ub3\_101, 2) = 0ub5\_00101 \\ 
{\operator{extend}}(0sb3\_101, 2) = 0sb5\_11101 \\ 
{\operator{extend}}(0sb3\_011, 2) = 0sb5\_00011 \\ 
\end{tabular}
\end{center}

\noindent Note that the right operand of \operator{extend} has to be 
an integer constant greater or equal to zero.

%-----    ------
\subsubsection{Resize Word Conversions}
\label{Resize Word Conversions}
\index{resize operator}

\operator{resize} operator provides a more comfortable way
of changing the width of a word. The behavior of this operator
can be described as follows:

let w be a M bits \UWord and N be the required width: if M = N, w is
returned unmodified; if N is less than M, bits in the range [N-1:0]
are extracted from w; if N is greater than M, w is extended of (N - M)
bits up to required width, padding with zeroes.

let w be a M bits \SWord and N be the required width: if M = N, w is
returned unmodified; if N is less than M, bits in the range [N-2:0]
are extracted from w, while N-1-ith bit is forced to preserve the
value of the original sign bit of w (M-1-ith bit); if N is greater than M, w is
extended of (N - M) bits up to required width, extending sign bit.

The signature of the operator is:\\

\begin{tabular}{l@{ : }l}
\operator{resize} 
& \UWord * \Integer $\rightarrow$ \UWord[\Integer]\\
& \SWord * \Integer $\rightarrow$ \SWord[\Integer]\\
\end{tabular}\\
%

%-----    ------
\subsubsection{Set Expressions}
\label{Set Expressions}
\index{set expressions}
\index{operators!union}

The set expression is an expression defining a set of \Boolean,
\Integer and \SymbEnum values. 
A set expression can be created with the \operator{union}
operator. For example, \code{1 \operator{union} 0} specifies the set
of values \code{1} and \code{0}. 
One or both of the operands of \operator{union} can be sets.  In this
case, \operator{union} returns a union of these sets. For example,
expression \code{(1 \operator{union} 0) \operator{union} -3} specifies
the set of values \code{1}, \code{0} and \code{-3}.

\emph{Note that there cannot be a set of sets in NuSMV}. 
Sets can contain only singleton values, but not other sets.

The signature of the \operator{union} operator is:\\
\begin{tabular}{ll}
\operator{union} 
 &{ : }\BoolSet * \BoolSet $\rightarrow$ \BoolSet\\
 &{ : }\IntSet * \IntSet $\rightarrow$ \IntSet\\
 &{ : }\SymbSet * \SymbSet $\rightarrow$ \SymbSet\\
 &{ : }\IntSymbSet * \IntSymbSet \\
 & \qquad $\rightarrow$ \IntSymbSet\\
\end{tabular}\\
Before checking the expression for being correctly typed, if it is
possible, both operands are converted to their counterpart \Set types
\footnote{See \ref{Set Types} for more information about the \Set types
and their counterpart types}, which virtually means converting
individual values to singleton sets. Then both operands are
implicitly converted to a minimal type that covers both operands.
If after these manipulations the operands do not satisfy the signature 
of \reserved{union} operator, an error is raised by \nusmv.


\index{expressions!sets}

There is also another way to write a set expression by enumerating all
its values between curly brackets. The syntactic rule for the values
in curly brackets is:\\
\begin{Grammar}
set_body_expr :: 
        basic_expr
      | set_body_expr \textbf{,} basic_expr
\end{Grammar}

Enumerating values in curly brackets is semantically equivalent to
writing them connected by \operator{union} operators. For example,
expression \code{\{exp1, exp2, exp3\}} is equivalent to \code{exp1
\operator{union} exp2 \operator{union} exp3}.  Note that according to
the semantics of \operator{union} operator, expression \code{\{\{1,
2\}, \{3, 4\}\}} is equivalent to \code{\{1, 2, 3, 4\}}, i.e.  there
is no actually set of sets.

 Set expressions can be used only as operands of \operator{union}
 and \operator{in} operations, as the right operand of \operator{case}
 and as the second and the third operand of \itebullet expressions and
 assignments.  In all other places the use of set expressions is
 prohibited.

%-----    ------
\subsubsection{Inclusion Operator \reserved{in}}
\label{Inclusion Operator}
\index{inclusion operator}
\index{operators!inclusion}

The inclusion operator `\reserved{in}' tests the left operand for
being a subset of the right operand. If either operand is a number or a
symbolic value instead of a set, it is coerced to a singleton set.

The signature of the \operator{in} operator is:\\

\begin{tabular}{ll}
\operator{in} &{ : }\BoolSet * \BoolSet $\rightarrow$ \Boolean\\
&{ : }\IntSet * \IntSet $\rightarrow$ \Boolean\\
&{ : }\SymbSet * \SymbSet $\rightarrow$ \Boolean\\
&{ : }\IntSymbSet * \IntSymbSet $\rightarrow$ \Boolean\\
\end{tabular}\\
Similar to \operator{union} operation, before checking the
expression for being correctly typed, if it is possible, both operands
are converted to their counterpart \Set types
\footnote{See \ref{Set Types} for more information about the \Set types
and their counterpart types}. Then, if required, implicit type
conversion is carried out on \emph{one} of the operands.


%-----    ------
\subsubsection{Case Expressions}
\label{Case Expressions}
\index{case expressions}
\index{expressions!case}

A case expression has the following syntax:
%
\begin{Grammar}
case_expr :: \reserved{case} case_body \reserved{esac}

case_body ::
        basic_expr \textbf{:} basic_expr {;}
      | case_body basic_expr \textbf{:} basic_expr {;}
\end{Grammar}

\noindent A \grammar{case\_expr} returns the value of the first
expression on the right hand side of `\code{:}', such that the
corresponding condition on the left hand side evaluates to
\code{TRUE}.
%
For example, the result of the expression
\begin{alltt}
\begin{tabular}{l}
\textbf{case}\\
 left_expression_1 \textbf{:} right_expression_1 \textbf{;}\\
 left_expression_2 \textbf{:} right_expression_2 \textbf{;}\\
 ...\\
 left_expression_N \textbf{:} right_expression_N \textbf{;}\\
\textbf{esac}
\end{tabular}
\end{alltt}

\noindent is \code{right\_expression\_k} such that for all $i$ from
$0$ to $k-1$, \code{left\_expression\_i} is \code{FALSE}, and
\code{left\_expression\_k} is \code{TRUE}.
%
It is an error if all expressions on the left hand side evaluate
to \code{FALSE}.

The type of expressions on the left hand side must be \Boolean.  If
one of the expression on the right is of a \Set type then, if it is
possible, all remaining expressions on the right are converted to
their counterpart \Set types \footnote{See \ref{Set Types} for
information on \Set types and their counterpart types}. The type of
the whole expression is such a minimal type\footnote{See \sref{Type
Order} for the information on the order of types.} that all of the
expressions on the right (after possible convertion to \Set types) can
be implicitly converted to this type.
%
If this is not possible, \nusmv throws an error.

\bigskip
\textbf{\textit{Note:}}
\label{ref::caseconditionexample}
Prior to version 2.5.1, using \code{1}
as \code{left\_expression\_N} was pretty common, e.g:
\begin{verbatim}
case 
  cond1 : expr1;
  cond2 : expr2;
  ...
  1     : exprN; -- otherwise
esac
\end{verbatim}

Since version 2.5.1 \Integer values are no longer implicitly casted
to \Boolean, and \code{1} has to be written as \code{TRUE} instead.
For backward compatibility options, please see page
\pageref{ref::backwardcompatibility}.

%-----    ------
\subsubsection{If-Then-Else expressions}
\label{If-Then-Else expressions}
\index{if-then-else expressions}

In certain cases, the syntax described above may look a bit awkard. In
simpler cases, it is possible to use the alternative, terser,
\itebullet expression. This construct is defined as follows:

\begin{Grammar}
        cond_expr \operator{?} basic_expr1 \operator{:} basic_expr2
\end{Grammar}

This expression evaluates to basic\_expr1 if the condition in
    cond\_expr evaluates to true, and to basic\_expr2 otherwise.
Therefore, the expressions \code{cond1 ? exp1 : exp2} and \code{case
    cond1 : exp1; \code{TRUE} : expr2; esac} are equivalent.

%-----    ------
\subsubsection{Basic Next Expression}
\label{Basic Next Expression}
\index{basic next expression}
\index{expressions!basic next}

\grammar{Next expressions} refer to the values of variables in the next
state.
%
For example, if a variable \code{\textbf{v}} is a state variable, then
\code{\textbf{next(v)}} refers to that variable \code{\textbf{v}} in
the next time step.
%
A \operator{next} applied to a complex expression is a
shorthand method of applying \operator{next} to all the variables in
the expressions recursively.
%
Example: \code{{\operator{next}}((1 + a) + b)} is equivalent to
\code{(1 + {\operator{next}}(a)) + {\operator{next}}(b)}.
%
Note that the \operator{next} operator cannot be applied twice, i.e.
\code{{\operator{next}}({\operator{next}}(a))} is \emph{not} allowed.

The syntactic rule is:

\begin{Grammar}
basic_next_expr :: \operator{next} \textbf{(} basic_expr \textbf{)}
\end{Grammar}
%
A \grammar{next expression} does not change the type.

%-----    ------
\subsubsection{Count Operator}
\label{count operator}
\index{operators!count}
The \operator{count} operator counts the number of expressions
which are true. 
%
The \operator{count} operator is a syntactic sugar for
\begin{Grammar}
     \operator{toint}\textbf{(}bool_expr1\textbf{)} +\\
     \operator{toint}\textbf{(}bool_expr2\textbf{)} +\\
     ... +\\
     \operator{toint}\textbf{(}bool_exprN\textbf{)}
\end{Grammar}
%
This operator has been introduced in version 2.5.1, to simplify
the porting of those models which exploited the implicit casting
of \Integer to \Boolean to encoding e.g. predicates like:

\code{ (b0 + b1 + ... + bN) < 3 -- at most two bits are enabled }
%
Since version 2.5.1, this expression can be written as:

\code{ count(b0 + b1 + ... + bN) < 3 }


%-----    ------
\subsection{Simple and Next Expressions}
\label{Simple and Next Expressions}
\index{simple expressions}
\index{next expressions}
\index{expressions!simple}
\index{expressions!next}

\grammar{Simple\_expressions} are expressions built only from the
values of variables in the current state.
%
Therefore, the \grammar{simple\_expression} cannot have a
\operator{next} operation inside and the syntax of
\grammar{simple\_ex\-pressions} is as follows:
%
\begin{Grammar}
simple_expr :: basic_expr
\end{Grammar}
%
with the alternative \grammar{basic\_next\_expr} \emph{not}
allowed. 
%
\grammar{Simple\_expressions} can be used to specify sets of states,
for example, the initial set of states.
%
The \grammar{next\_expression} relates current and next state
variables to express transitions in the FSM. 
%
The \grammar{next\_expression} \emph{can} have \operator{next}
operation inside, i.e.
%
\begin{Grammar}
next_expr :: basic_expr
\end{Grammar}
%
with the alternative \grammar{basic\_next\_expr} allowed.
%
%\index{expressions!next set}
%\index{expressions!simple set}
%%
%There are also similar expressions to denote
%\grammar{set\_expressions}, i.e.
%%
%\begin{Grammar}
%%next_expr :: basic_expr
%simple_expr :: basic_expr
%\end{Grammar}
%
%where \grammar{next\_expr} may have any expressions defined by
%\grammar{basic\_expr}, and \grammar{simple\_expr} cannot
%have the keyword \operator{next} inside.

%-----    ------
\subsection{Type conversion operators}
\label{Type conversion operators}
\index{Type conversion operators}
\index{operators!cast}


%-----    ------
\subsubsection{Integer conversion operator}
\label{Integer Conversion Operator}
\index{toint operator}

\operator{toint} converts an \UWord \grammar{constant} or a 
\SWord \grammar{constant}, or a \Boolean expression to an 
\Integer representing its value. Also \Integer expressions are 
allowed, but no action is performed.
%
The signature of this conversion operator is:\\

\begin{tabular}{l@{ : }l}
\operator{toint} & \Integer $\rightarrow$ \Integer\\
\operator{toint} & \Boolean $\rightarrow$ \Integer\\
\operator{toint} & \UWord $\rightarrow$ \Integer\\
\operator{toint} & \SWord $\rightarrow$ \Integer\\
\end{tabular}\\
%
\\
Warning: using the \operator{toint} operator with word variables may
cause bad performances of the system. Performances may degrade with
the increase of the number of bits of the word expression.

%-----    ------
\subsubsection{Boolean conversion operator}
\label{Boolean Conversion Operator}
\index{bool operator}

\operator{bool} converts \UWord[1] and \Integer expressions
to \Boolean. Also \Boolean expressions are allowed, but no 
action is perfomed.
%
In case of \Integer expression, the result of the conversion
is \code{FALSE} if the expression resolves to \code{0},
\code{TRUE} otherwise.  In case of \UWord[1] expression, the conversion obeys
the following table:
%
\begin{center}
\begin{tabular}{p{0.3\textwidth}p{0.3\textwidth}}
{\operator{bool}}(0ub1\_0) = FALSE\\ 
{\operator{bool}}(0ub1\_1) = TRUE\\
\end{tabular}
\end{center}

\subsubsection{Integer to Word Constants Conversion}
\label{Integer to Word Constants Conversion}
\index{swconst operator}
\index{uwconst operator}

\operator{swconst}, \operator{uwconst} convert an \Integer \grammar{constant}
into a \SWord \grammar{constant} or \UWord \grammar {constant} of
given size respectively.
%
The signature of these conversion operator is:\\

\begin{tabular}{l@{ : }l}
\operator{swconst} & \Integer * \Integer $\rightarrow$ \SWord\\
\operator{uwconst} & \Integer * \Integer $\rightarrow$ \UWord\\
\end{tabular}\\

\noindent Where the left \Integer parameter is the \textbf{value}
and the right \Integer parameter is the \textbf{size} in bits of the
generated \UWord or \SWord \grammar{constant}.

\subsubsection{Word1 Explicit Conversions}
\label{Word[1] Explicit Conversions}
\index{word1 operator}

\operator{word1} converts a \Boolean to a \UWord[1].
%
The signature of this conversion operator is:\\

\begin{tabular}{l@{ : }l}
\operator{word1} & \Boolean $\rightarrow$ \UWord[1]\\
\end{tabular}\\

\noindent The conversion obeys the following table:
%
\begin{center}
\begin{tabular}{p{0.3\textwidth}p{0.3\textwidth}}
{\operator{word1}}(FALSE) = 0ub1\_0\\ 
{\operator{word1}}(TRUE) = 0ub1\_1\\
\end{tabular}
\end{center}

%-----    ------
\subsubsection{Unsigned and Signed Explicit Conversions}
\label{Unsigned and Signed Explicit Conversions}
\index{signed operator}
\index{unsigned operator}

\operator{unsigned} converts a \SWord[N] to an \UWord[N], while
\operator{signed} performs the opposite operation and converts 
an \UWord[N] to a \SWord[N]. Both operations do not change the bit
representation of a provided word.
%
The signatures of these conversion operators are:\\

\begin{tabular}{l@{ : }l}
\operator{unsigned} & \SWord[N] $\rightarrow$ \UWord[N]\\
\operator{signed} & \UWord[N] $\rightarrow$ \SWord[N]\\
\end{tabular}\\

\noindent For example:
%
\begin{center}
\begin{tabular}{l}
{\operator{signed}}(0ub\_101) = 0sb\_101 \\ 
{\operator{signed}}(0ud3\_5) = -0sd3\_3 \\ 
{\operator{unsigned}}(0sb\_101) = 0usb\_101 \\ 
{\operator{unsigned}}(-0sd3\_3) = 0ud3\_5 \\ 
\end{tabular}
\end{center}




%-------------------- DEFINITION OF FSM -----------------------------------

\section{Definition of the FSM}
\label{Definition of the FSM}
\index{definition of the FSM}

We consider a Finite State Machine (FSM) described in terms of
\emph{state variables}, \emph{input variables}, and \emph{frozen
variables}, which may assume different values in different
\emph{states}, of a \emph{transition relation} describing how inputs
leads from one state to possibly many different states, and of
\emph{Fairness conditions} that describe constraints on the valid
paths of the execution of the FSM.
%
In this document, we distinguish among constraints (used to constrain
the behavior of a FSM, e.g.\ a modulo 4 counter increments its value
modulo 4), and specifications (used to express properties to verify on
the FSM (e.g.\ the counter reaches value 3).

In the following it is described how these concepts can be declared in
the \nusmv language.

%-----    ------
\subsection{Variable Declarations}
\label{Variable Declarations}
\index{variable declarations}

A variable can be an input, a frozen, or a state variable. 
%
The declaration of a variable specifies the variable's type with the
help of type specifier.

%-----    ------
\subsubsection{Type Specifiers}
\label{Type Specifiers}
\index{type specifiers}
%
\index{syntax rules!type specifiers}
A \grammar{type specifier} has the following syntax:
%
\begin{Grammar}
type_specifier ::
        simple_type_specifier
      | module_type_specifier

simple_type_specifier :: 
        \textbf{boolean}
      | \textbf{word} \textbf{[} basic_expr \textbf{]} 
      | \textbf{unsigned word} \textbf{[} basic_expr \textbf{]} 
      | \textbf{signed word} \textbf{[} basic_expr \textbf{]} 
      | \textbf{\{} enumeration_type_body \textbf{\}}
      | basic_expr \textbf{..} basic_expr
      | \textbf{array} basic_expr \textbf{..} basic_expr
        \ \ \textbf{of} simple_type_specifier

enumeration_type_body ::
        enumeration_type_value
      | enumeration_type_body \textbf{,} enumeration_type_value

enumeration_type_value ::
        symbolic_constant
      | integer_number
\end{Grammar}
% 
There are two kinds of \grammar{type specifier}: a \grammar{simple
type specifier} and a \grammar{module type specifier}. 
%
The \grammar{module type specifier} is explained later in
\sref{MODULE Instantiations}.
%
The \grammar{simple type specifier} comprises \Boolean type, \Integer
type, \Enum types, \UWord, \SWord and arrays 
%and \WordArray[] 
types.

The \Boolean type is specified by the keyword \reserved{boolean}. 
%The \Integer type is specified by the keyword \reserved{Integer}. 

A \Enum type is specified by full enumeration of all the values the
type comprises. 
%
For example, possible \Enum type specifiers are \code{\{0,2,3,-1\}},
\code{\{1,0, OK\}}, \code{\{OK, FAIL, running\}}. \code{FALSE} and 
\code{TRUE} values cannot be used as \Enum type specifiers
%
The values in the list are enclosed in curly brackets and separated by
commas.
%
The values may be \grammar{integer numbers}, \grammar{symbolic
constants}, or both.
%
All values in the list should be distinct from each other, although the
order of values is not important. 

Note, expressions cannot be of the actual \Enum types, but only the
simplified versions of \Enum types, such as \SymbEnum and \IntSymbEnum.

A \grammar{type specifier} can be given by two expressions
separated by \code{\textbf{..}} (\code{<TWO DOTS>}). The two
expressions have both to evaluate to constants integer numbers, and
may contain names of defines and module formal parameters. For example,
\code{-1 - P1 .. 5 + D1}, where \code{P1} refers to a module formal 
parameter, and \code{D1} refers to a define. Both \code{P1} and 
\code{D1} have to be statically evaluable to integer constants.

This is just a shorthand for a \Enum type containing the list
of \grammar{integer numbers} from the range given in \grammar{type
specifier}.
%
For example, the \grammar{type specifiers} \code{-1..5} and
\code{\{-1,0,1,2,3,4,5\}} are equivalent. 
%
Note that the evaluated number on the left from the two dots must
be less than or equal to the evaluated number on the right.

The \UWord type is specified by the keywords \reserved{unsigned word}
(where \reserved{unsigned} may be skipped) with a \grammar{basic\_expr} 
supplied in square brackets.
%
The expression must be statically evaluable to a constant integer
number whose value must be greater than zero. 
%
The \SWord type is specified in a similar way with the keywords
\reserved{signed word}.
%
The purpose of the word types is to offer integer and bitwise
arithmetic.


An array type is denoted by a sequence of the keyword
\reserved{array}, a \grammar{basic\_expr} specifying the lower
bound of the array index, two dots \code{\textbf{..}}, a
\grammar{basic\_expr} specifying the upper bound of the array
index, the keyword \reserved{of}, and the type of array's
elements. The elements can themselves be arrays. The two bound
expressions have to be statically evaluable to constant integer
numbers, and may contain names of defines and module formal
parameters.


%-----    ------
\subsubsection{State Variables}
\label{State Variables}
\index{state variables}
\index{\code{VAR} declaration}
%
A state of the model is an assignment of values to a set of state
and frozen variables. 
%
State variables (and also instances of modules) are declared by the
notation:
%
\begin{Grammar}
var_declaration :: \reserved{VAR} var_list

var_list :: identifier \textbf{:} type_specifier \textbf{;}
          | var_list identifier \textbf{:} type_specifier \textbf{;}
\end{Grammar}
%
A \grammar{variable declaration} specifies the identifier of the
variables and its type. 
%
A variable can take the values only from the domain of its type. In
particular, a variable of a \Enum type may take only the values
enumerated in the \grammar{type specifier} of the declaration.

%-----    ------
\subsubsection{Input Variables}
\label{Input Variables}
\index{input variables syntax}
\index{\code{IVAR} declaration}
%
\code{IVAR}s (input variables) are used to label transitions of the
Finite State Machine. The difference between the syntax for the input
and state variables declarations is the keyword indicating the
beginning of a declaration:
%
\begin{Grammar}
ivar_declaration :: \reserved{IVAR} simple_var_list
simple_var_list ::
        identifier \textbf{:} simple_type_specifier \textbf{;}
      | simple_var_list identifier \textbf{:} simple_type_specifier \textbf{;}
\end{Grammar}
%
Another difference between input and state variables is that input
variables cannot be instances of modules.
%
The usage of input variables is more limited than the usage of state
variables which can occur everywhere both in the model and
specifications.
%
Namely, input variables cannot occur in:

\begin{itemize}
\item Left-side of assignments. For example all these assignments 
  are not allowed:

  \code{IVAR i : boolean;}\\
  \code{ASSIGN}\\
  \code{init(i) := TRUE;}\\
  \code{next(i) := FALSE;}

\item \code{INIT} statements. For example:

  \code{IVAR i : boolean;}\\
  \code{VAR s : boolean;}\\
  \code{INIT i = s}

\item Scope of \code{next} expressions. For example:

  \code{IVAR i : boolean;}\\
  \code{VAR s : boolean;}\\
  \code{TRANS i -> s} -- this is allowed\\
  \code{TRANS next(i -> s)} -- this is NOT allowed

\item Some specification kinds: \code{CTLSPEC}, \code{SPEC},
      \code{INVARSPEC}, \code{COMPUTE}, \code{PSLSPEC}. For example:

  \code{IVAR i : boolean;}\\
  \code{VAR s : boolean;}\\
  \code{SPEC AF (i -> s)} -- this is NOT allowed\\
  \code{LTLSPEC F (X i -> s)} -- this is allowed


\item Anywhere in the FSM when checking invariants with BMC and the
  ``DUAL'' algorithm. See at page \pageref{bmc::dual} for further
  information.

\end{itemize}


%-----    ------
\subsubsection{Frozen Variables}
\label{Frozen Variables}
\index{frozen variables syntax}
\index{\code{FROZENVAR} declaration}
%
\code{FROZENVAR}s (frozen variables) are variables that retain their
initial value throughout the evolution of the state machine; this
initial value can be constrained in the same ways as for normal state
variables. Similar to input variables the difference between the
syntax for the frozen and state variables declarations is the keyword
indicating the beginning of a declaration:
%
\begin{Grammar}
frozenvar_declaration :: \reserved{FROZENVAR} simple_var_list
\end{Grammar}
%
The semantics of some frozen variable \code{a} is that of a state
variable accompanied by an assignment that keeps its value constant
(it is handled more efficiently, though):

\begin{tabular}{@{\hspace{1cm}}l}\\
\code{ASSIGN next(a) := a;}\\\\
\end{tabular}

As a consequence, frozen variables may not have their current and next
value set in an \code{ASSIGN} statement, i.e. statements such as
\code{ASSIGN next(a) := expr;} and \code{ASSIGN a := expr;} are
illegal. Apart from that frozen variables may occur in the definition
of the FSM in any place in which a state variable may occur. Some
examples are as follows:

\begin{itemize}

\item Left-side current and next state assignments are illegal, while
init state assignments are allowed:

  \code{FROZENVAR a :~boolean;}\\
  \code{FROZENVAR b :~boolean;}\\
  \code{FROZENVAR c :~boolean;}\\
  \code{VAR d :~boolean;}\\
  \code{FROZENVAR e :~boolean;}\\
  \code{ASSIGN}\\
  \code{init(a) := d; -- legal}\\
  \code{next(b) := d; -- illegal}\\
  \code{c       := d; -- illegal}\\
  \code{e       := a; -- also illegal}

\item \code{INIT}, \code{TRANS}, \code{INVAR}, \code{FAIRNESS},
\code{JUSTICE}, and \code{COMPASSION} statements are all legal. So is
the scope of a \code{next} expression. For example:

  \code{-- the following has an empty state space}\\
  \code{FROZENVAR a :~boolean;}\\
  \code{INIT a}\\
  \code{INVAR !a}\\
  \code{}\\
  \code{-- alternatively, this has two initial states, deadlocking}\\
  \code{FROZENVAR b :~boolean;}\\
  \code{TRANS next(b) <-> !b}\\
  \code{}\\
  \code{-- and that's just unfair}\\
  \code{FROZENVAR c :~boolean;}\\
  \code{FAIRNESS c}\\
  \code{FAIRNESS !c}

\item All kinds of specifications involving frozen variables are
allowed, e.g.:

  \code{FROZENVAR c :~boolean;}\\
  \code{-- True by definition.}\\
  \code{SPEC AG ((c -> AG c) \& ((!c) -> AG !c))}\\
  \code{-- Here, neither is true.}\\
  \code{INVARSPEC c}\\
  \code{INVARSPEC !c}\\
  \code{-- False (as above).}\\
  \code{LTLSPEC (G F c) \& (G F !c)}

\end{itemize}

%-----    ------
\subsubsection{Examples}
\label{Examples of variable declaration}
\index{input variables syntax}
\index{state variables syntax}
\index{\code{IVAR} declaration}
\index{\code{VAR} declaration}

Below are examples of state, frozen, and input variable declarations:

\begin{tabular}{@{\hspace{1cm}}l}\\
\code{VAR a :~boolean;}\\
\code{FROZENVAR b :~0..1;}\\
\code{IVAR c :~\{TRUE, FALSE\};}\\\\
\end{tabular}

\noindent The variable \code{a} is a state variable,
\code{b} is a frozen variable, and \code{c} is an input variable;
%
In the following examples:

\begin{tabular}{@{\hspace{1cm}}l}\\
\code{VAR d :~\{stopped, running, waiting, finished\};}\\
\code{VAR e :~\{2, 4, -2, 0\};}\\
\code{VAR f :~\{1, a, 3, d, q, 4\};}\\\\
\end{tabular}

\noindent the variables \code{d}, \code{e} and \code{f} are of \Enum types,
and all their possible values are specified in the \grammar{type
specifiers} of their declarations.

\begin{tabular}{@{\hspace{1cm}}l}\\
\code{VAR g :~unsigned word[3];}\\\\
\code{VAR h :~word[3];}\\\\
\code{VAR i :~signed word[4];}\\\\
\end{tabular}

\noindent The variables \code{g} and \code{h} are of 3-bits-wide
\type{unsigned word} type (i.e.\@ \UWord[3]), and \code{i} is of
4-bits-wide \type{signed word} type (i.e.\@ \SWord[4]).

\begin{tabular}{@{\hspace{1cm}}l}\\
\code{VAR j :~array -1..1 of boolean;}\\\\
\end{tabular}

\noindent The variable \code{j} is an array of \Boolean elements with
indexes -1, 0 and 1.

% For examples of module instantiation see \sref{MODULE Instantiations}.

%-----    ------
% \code{} used instead of \reserved{} so that font size is not reduced
\subsection{\code{DEFINE} Declarations}
\label{DEFINE Declarations}
\index{DEFINE declarations}
%
In order to make descriptions more concise, a symbol can be associated
with a common expression, and a \reserved{DEFINE} declaration
introduces such a symbol.
%
The syntax for this kind of declaration is:
%
\begin{Grammar}
define_declaration :: \reserved{DEFINE} define_body

define_body :: identifier \operator{:=} simple_expr \textbf{;}
             | define_body identifier \operator{:=} simple_expr \textbf{;}
\end{Grammar}
%
\reserved{DEFINE} associates an \grammar{identifier} on the left hand
side of the \operator{`:='} with an expression on the right side.
%
A define statement can be considered as a macro. 
%
Whenever a define \grammar{identifier} occurs in an expression, the
\grammar{identifier} is syntactically replaced by the expression it is
associated with.
%
The associated expression is always evaluated in the context of the
statement where the \grammar{identifier} is declared (see \sref{Context}
for an explanation of contexts).
%
Forward references to defined symbols are allowed but circular
definitions are not, and result in an error.
%
The difference between defined symbols and variables is that while
variables are statically typed, definitions are not.

%-----  ARRAY DEFINE DECLARATION  ------
\subsection{Array Define Declarations}
\label{Array Define Declarations}
\index{array define declarations}
\index{DEFINE : array}
%
It is possible to specify an array expressions. This
feature is experimental and currently available only through
\reserved{DEFINE} declaration.
%
The syntax for this kind of declaration is:
%
\begin{Grammar}
array_define_declaration :: 
                  \reserved{DEFINE} identifier \operator{:=} array_expression \textbf{;}

array_expression :: \operator{[} array_contents \operator{]}
                   | \operator{[} array_expression_list \operator{]}

array_expression_list :: array_expression
                        | array_expression \operator{,} array_expression_list

array_contents :: next_expr \operator{,} array_contents
                 | next_expr
\end{Grammar}
%
Array \reserved{DEFINE} associates an \grammar{identifier} on the left hand
side of the \operator{`:='} with an array expression.
%
As a normal \reserved{DEFINE} statement an array define is considered
as a macro.
%
Whenever an array \grammar{identifier} occurs in an expression, the
\grammar{identifier} is syntactically replaced by the array expression it is
associated with.
%
As with normal \reserved{DEFINE} an array \reserved{DEFINE} expression
is always evaluated in the context of the statement where the
\grammar{identifier} is declared and forward references to defined symbols are
allowed but circular definitions are not. 

The type of an array expression \code{[exp1, exp2, \ldots, expN]} is
\Array 0..N-1 of \code{type} where \code{type} is the least type such that 
all \code{exp1, exp2, \ldots expN} can be converted to it.


It is not possible to declare asymmetrical arrays. This means that
it is forbidden to declare an array with a different number of
elements in a dimension. For example, the following code will result
in an error:
%
\begin{nusmvCode}
DEFINE
   x := [[1,2,3], [1,2]];
\end{nusmvCode}

%-----    ------
\subsection{\code{CONSTANTS} Declarations}
\label{CONSTANTS Declarations}
\index{CONSTANTS declarations}
%
\reserved{CONSTANTS} declarations allow the user to explicitly 
declare symbolic constants that might occur or not within the FSM that
is being defined. 
%
\reserved{CONSTANTS} declarations are expecially useful 
in those conditions that require symbolic constants to occur only
in \reserved{DEFINEs} body (e.g. in generated models).
For an example of usage see also the command \command{write\_boolean\_model}.
%
A constant is allowed to be declared multiple times, as after the
first declaration any further declaration will be ignored.
%
\reserved{CONSTANTS} declarations are an extension of the original 
\smv grammar, and they are supported since \NuSMV.
%
The syntax for this kind of declaration is:
%
\begin{Grammar}
constants_declaration :: \reserved{CONSTANTS} constants_body \textbf{;}

constants_body :: identifier 
             | constants_body  \operator{,} identifier
\end{Grammar}


%-----    ------
\subsection{\code{INIT} Constraint}
\label{INIT Constraint}
\index{INIT constraint}
%
The set of initial states of the model is determined by a \Boolean
expression under the \reserved{INIT} keyword.
%
The syntax of an \code{INIT} constraint is:
%
\begin{Grammar}
init_constrain :: \reserved{INIT} simple_expr [\textbf{;}]
\end{Grammar}
%
Since the expression in the \code{INIT} constraint is a
\grammar{simple\_expression}, it cannot contain the \operator{next()}
operator.
%
The expression also has to be of type \Boolean.
%
If there is more than one \code{INIT} constraint, the initial set is
the conjunction of all of the \code{INIT} constraints.

%-----    ------
\subsection{\code{INVAR} Constraint}
\label{INVAR Constraint}
\index{INVAR constraint}
%
The set of invariant states can be specified using a \Boolean
expression under the \reserved{INVAR} keyword.
%
The syntax of an \code{INVAR} constraint is:
%
\begin{Grammar}
invar_constraint :: \reserved{INVAR} simple_expr [\textbf{;}]
\end{Grammar}
%
Since the expression in the \code{INVAR} constraint is a
\grammar{simple\_expression}, it cannot contain the \operator{next()}
operator. 
%
If there is more than one \code{INVAR} constraint, the invariant set
is the conjunction of all of the \code{INVAR} constraints.

%-----    ------
\subsection{\code{TRANS} Constraint}
\label{TRANS Constraint}
\index{TRANS constraint}
%
The transition relation of the model is a set of current state/next
state pairs. 
%
Whether or not a given pair is in this set is determined by a boolean
expression, introduced by the \reserved{TRANS} keyword. 
%
The syntax of a \code{TRANS} constraint is:
%
\begin{Grammar}
trans_constraint :: \reserved{TRANS} next_expr [\textbf{;}]
\end{Grammar}
%
It is an error for the expression to be not of the \Boolean type.
%
If there is more than one \code{TRANS} constraint, the transition
relation is the conjunction of all of \code{TRANS} constraints.

% \code{} used instead of \reserved{} so that font size is not reduced
\subsection{\code{ASSIGN} Constraint}
\label{ASSIGN Constraint}
\index{\code{ASSIGN} constraint}
%
An assignment  has the form:
%
\begin{Grammar}
assign_constraint :: \reserved{ASSIGN} assign_list

assign_list :: assign \textbf{;}
             | assign_list assign \textbf{;}

assign ::
    complex_identifier          \operator{:=} simple_expr
  | \reserved{init} \textbf{(} complex_identifier \textbf{)} \operator{:=} simple_expr
  | \reserved{next} \textbf{(} complex_identifier \textbf{)} \operator{:=} next_expr
\end{Grammar}
%
On the left hand side of the assignment, \grammar{identifier} denotes
the current value of a variable,
`{\reserved{init}}\code{\textbf{(}identifier\textbf{)}}' denotes its
initial value, and
`{\reserved{next}}\code{\textbf{(}identifier\textbf{)}}' denotes its
value in the next state.
%
If the expression on the right hand side evaluates to a not-\Set
expression such as \grammar{integer number} or \grammar{symbolic
constant}, the assignment simply means that the left hand side is
equal to the right hand side.
%
On the other hand, if the expression evaluates to a set, then the
assignment means that the left hand side is contained in that set. 
%
It is an error if the value of the expression is not contained in the
range of the variable on the left hand side.

Semantically assignments can be expressed using other kinds of
constraints:\\

\begin{tabular}{l@{\ is equivalent to\  }l}
%\multicolumn{2}{c}{\ }\\
\code{ASSIGN a := exp;} & \code{INVAR a in exp;}\\
\code{ASSIGN init(a) := exp;} & \code{INIT a in exp;}\\
\code{ASSIGN next(a) := exp;} & \code{TRANS next(a) in exp;}\\
\multicolumn{2}{c}{\ }\\
\end{tabular}\\

\noindent Notice that, an additional constraint is forced when
assignments are used with respect to their corresponding constraints
counterpart: when a variable is assigned a value that it is not an
element of its declared type, an error is raised.

The allowed types of the assignment operator are:

\begin{tabular}{ll}
\operator{:=} &{ : }\Integer * \Integer\\
&{ : }\Integer * \IntSet \\
&{ : }\SymbEnum * \SymbEnum\\
&{ : }\SymbEnum * \SymbSet\\
&{ : }\IntSymbEnum * \IntSymbEnum \\
&{ : }\IntSymbEnum * \IntSymbSet \\
&{ : }\UWord[N] * \UWord[N] \\
&{ : }\SWord[N] * \SWord[N] \\
% &{ : }\WordArray[{[N][M]}] * \WordArray[{[N][M]}] \\
\end{tabular}\\
Before checking the assignment for being correctly typed,
the implicit type conversion can be applied to the \emph{right} operand.

%-----    ------
\subsubsection{Rules for assignments}

Assignments describe a system of equations that say how the FSM
evolves through time.
%
With an arbitrary set of equations there is no guarantee that a
solution exists or that it is unique.
%
We tackle this problem by placing certain restrictive syntactic rules
on the structure of assignments, thus guaranteeing that the program is
implementable.

The restriction rules for assignments are:
%
\begin{itemize}
  \item \textbf{The single assignment rule} -- each variable may be
        assigned only once.


  \item \textbf{The circular dependency rule} -- a set of equations
        must not have ``cycles'' in its dependency graph not broken by delays.
\end{itemize}

The single assignment rule disregards conflicting definitions, and can
be formulated as: one may either assign a value to a variable
``\code{x}'', or to ``\reserved{next(}x\reserved{)}'' and
``\reserved{init(}x\reserved{)}'', but not both. For instance, the
following are legal assignments:

\begin{center}
  \begin{tabular}{|l|l|}\hline
    Example 1 & \code{x} \reserved{:=} \code{expr$_1$}\reserved{;} \\
    \ & \ \\\hline
    Example 2 & \reserved{init(}\code{x}\reserved{)} \reserved{:=} \code{expr$_1$}\reserved{;} \\
    \ & \ \\\hline
    Example 3 & \reserved{next(}\code{x}\reserved{)} \reserved{:=} \code{expr$_1$}\reserved{;} \\
    \ & \ \\\hline
    Example 4 & \reserved{init(}\code{x}\reserved{)} \reserved{:=} \code{expr$_1$}\reserved{;} \\
              & \reserved{next(}\code{x}\reserved{)} \reserved{:=} \code{expr$_2$}\reserved{;} \\
    \hline
\end{tabular}
\end{center}
while the following are illegal assignments:

\begin{center}
  \begin{tabular}{|l|l|}\hline
    Example 1 & \code{x} \reserved{:=} \code{expr$_1$}\reserved{;} \\
    \ & \code{x} \reserved{:=} \code{expr$_2$}\reserved{;} \\\hline
    Example 2 & \reserved{init(}\code{x}\reserved{)} \reserved{:=} \code{expr$_1$}\reserved{;} \\
    \ & \reserved{init(}\code{x}\reserved{)} \reserved{:=} \code{expr$_2$}\reserved{;} \\\hline
    Example 3 & \code{x} \reserved{:=} \code{expr$_1$}\reserved{;} \\
    \ & \reserved{init(}\code{x}\reserved{)} \reserved{:=} \code{expr$_2$}\reserved{;}\\\hline
    Example 4 & \code{x} \reserved{:=} \code{expr$_1$}\reserved{;} \\
              & \reserved{next(}\code{x}\reserved{)} \reserved{:=} \code{expr$_2$}\reserved{;} \\
    \hline
\end{tabular}
\end{center}

If we have an assignment like \code{x} \reserved{:=} \code{y}
\reserved{;}, then we say that \code{x} \emph{depends on} \code{y}. A
\emph{combinatorial loop} is a cycle of dependencies not broken by
delays. For instance, the assignments:
%
\begin{nusmvCode}
x := y;
y := x;
\end{nusmvCode}
%
form a combinatorial loop. Indeed, there is no fixed order in which we
can compute \code{x} and \code{y}, since at each time instant the
value of \code{x} depends on the value of \code{y} and vice-versa. We
can introduce a ``unit delay dependency'' using the \reserved{next()}
operator.
%
\begin{nusmvCode}
      x := y;
next(y) := x;
\end{nusmvCode}
%
In this case, there is a unit delay dependency between \code{x} and
\code{y}. 
%
A combinatorial loop is a cycle of dependencies whose total delay is
zero. In \nusmv combinatorial loops are illegal. This guarantees that
for any set of equations describing the behavior of variable, there is
at least one solution. There might be multiple solutions in the case
of unassigned variables or in the case of non-deterministic
assignments such as in the following example,
%
\begin{nusmvCode}
    next(x) := case x = 1 : 1;
                    TRUE  : \{0,1\};
               esac;
\end{nusmvCode}



%-----    ------
\subsection{\code{FAIRNESS} Constraints}
\label{FAIRNESS Constraints}
\index{\code{FAIRNESS} constraints}
\index{fairness constraints}
\index{justice constraints}
\index{compassion constraints}
\index{fair paths}
%
A fairness constraint restricts the attention only to \dfn{fair
execution paths}. When evaluating specifications, the model checker
considers path quantifiers to apply only to fair paths.

\nusmv supports two types of fairness constraints, namely justice
constraints and compassion constraints. 
%
A justice constraint consists of a formula \code{f}, which is assumed
to be true infinitely often in all the fair paths.
%
In \nusmv, justice constraints are identified by
keywords \reserved{JUSTICE} and, for backward compatibility,
\reserved{FAIRNESS}.
%
A compassion constraint consists of a pair of formulas \code{(p,q)};
if property \code{p} is true infinitely often in a fair path, then
also formula \code{q} has to be true infinitely often in the fair
path.
%
In \nusmv, compassion constraints are identified by keyword
\reserved{COMPASSION}.~\footnote{In the current version of \nusmv,
compassion constraints are supported only for BDD-based LTL model
checking. We plan to add support for compassion constraints also for
CTL specifications and in Bounded Model Checking in the next releases
of \nusmv.} 
%
If compassion constraints are used, then the model must not contain any
input variables. 
%
Currently, \nusmv does not enforce this so it is the responsibility of
the user to make sure that this is the case.

Fairness constraints are declared using the following syntax (all
expressions are expected to be \Boolean):
%
\begin{Grammar}
fairness_constraint ::
       \reserved{FAIRNESS} simple_expr [\textbf{;}]
     | \reserved{JUSTICE} simple_expr [\textbf{;}]
     | \reserved{COMPASSION} \textbf{(} simple_expr \textbf{,} simple_expr \textbf{)} [\textbf{;}]
\end{Grammar}
%
A path is considered fair if and only if it satisfies all the
constraints declared in this manner.


%-----    ------
\subsection {\code{MODULE} Declarations}
\label{MODULE Declarations}
\index{MODULE declarations}
%
A module declaration is an encapsulated collection of declarations,
constraints and specifications. 
%
A module declaration also opens a new
identifier scope.
%
Once defined, a module can be reused as many times as
necessary. 
%
Modules are used in such a way that each instance of a module refers to
different data structures. 
%
A module can contain instances of other modules, allowing a structural
hierarchy to be built. 
%
The syntax of a module declaration is as follows:
%
\index{syntax rules!module declarations}
\begin{Grammar}
module :: \reserved{MODULE} identifier [\textbf{(} module_parameters \textbf{)}] [module_body]

module_parameters ::
          identifier
        | module_parameters \textbf{,} identifier

module_body :: 
          module_element 
        | module_body module_element
           
module_element ::
          var_declaration
        | ivar_declaration
        | frozenvar_declaration
        | define_declaration
        | constants_declaration
        | assign_constraint
        | trans_constraint
        | init_constraint
        | invar_constraint
        | fairness_constraint
        | ctl_specification
        | invar_specification
        | ltl_specification
        | compute_specification
        | isa_declaration
\end{Grammar}
%
The \grammar{identifier} immediately following the keyword
\reserved{MODULE} is the name associated with the module. 
%
Module names have a separate name space in the program, and hence may
clash with names of variables and definitions. 
%
The optional list of identifiers in parentheses are the formal
parameters of the module.

%-----    ------
\subsection {\code{MODULE} Instantiations}
\label{MODULE Instantiations}
\index{MODULE instantiations}

An \emph{instance} of a module is created using the \reserved{VAR}
declaration (see \sref{State Variables}) with a module type specifier
(see \sref{Type Specifiers}).
%
The syntax of a \grammar{module type specifier} is:

\begin{Grammar}
module_type_specifier ::     
      | identifier [ \textbf{(} [ parameter_list ] \textbf{)} ]
      | \reserved{process} identifier [ \textbf{(} [ parameter_list ] \textbf{)} ]

parameter_list ::
        next_expr
      | parameter_list \textbf{,} next_expr
\end{Grammar}
%
A variable declaration with a \grammar{module type specifier}
introduces a name for the module instance. 
%
The \grammar{module type specifier} provides the name of the
instantiating module and also a list of actual parameters, which are
assigned to the formal parameters of the module.
%
An actual parameter can be any legal \grammar{next expression} (see
\sref{Simple and Next Expressions}). 
%
It is an error if the number of actual parameters is different from
the number of formal parameters. 
%
Whenever formal parameters occur in expressions within the module,
they are replaced by the actual parameters. 
%
The semantic of module instantiation is similar to
call-by-reference.\footnote{This also means that the actual parameters
are analyzed in the context of the variable declaration where the module
is instantiated, not in the context of the expression where the formal
parameter occurs.}

Here are examples:
%
\begin{nusmvCode}
MODULE main
...
 VAR
  a : boolean;
  b : foo(a);
...
MODULE foo(x)
 ASSIGN
   x := TRUE;
\end{nusmvCode}
%
the variable \code{a} is assigned the value \code{TRUE}. 
%
This distinguishes the call-by-reference mechanism from a
call-by-value scheme.

\noindent Now consider the following program:
%
\begin{nusmvCode}
MODULE main
...
 DEFINE
   a := 0;
 VAR
   b : bar(a);
...
MODULE bar(x)
 DEFINE
   a := 1;
   y := x;
\end{nusmvCode}
%
In this program, the value of \code{y} is \code{0}. On the other hand,
using a call-by-name mechanism, the value of \code{y} would be
\code{1}, since \code{a} would be substituted as an expression for
\code{x}.

\noindent Forward references to module names are allowed, but circular
references are not, and result in an error.

The keyword \reserved{process} is explained in \sref{Processes}.

%-----    ------
\subsection{References to Module Components (Variables and Defines)}
\label{References to Module Components}
\index{identifiers}

As described in \sref{Variables and Defines}, defines and variables
can be referenced in expressions as \code{variable\_identifiers} and
\code{define\_identifiers} respectively, both of which are
\grammar{complex identifiers}. 
%
The syntax of a \grammar{complex identifier} is:\\

\index{syntax rules!complex identifiers}
\begin{Grammar}
complex_identifier ::
        identifier
      | complex_identifier \textbf{.} identifier
      | complex_identifier \textbf{[} simple_expression \textbf{]}
      | \reserved{self}
\end{Grammar}

Every variable and define used in an expression should be declared. It
is possible to have forward references when a variable or define
identifier is used textually before the corresponding declaration.

Notations with \code{\textbf{.}} (\code{<DOT>}) are used to access the
components of modules.
%
For example, if \code{m} is an instance of a module (see \sref{MODULE
Instantiations} for information about instances of modules) then the
expression \code{m.c} identifies the component \code{c} of the module
instance \code{m}. 
%
This is precisely analogous to accessing a component of a structured
data type.

Note that actual parameters of a module can potentially be instances
of other modules.
%
Therefore, parameters of modules allow access to the components of
other module instances, as in the following example:
%
\begin{nusmvCode}
MODULE main
...  VAR
  a : bar;
  m : foo(a);
...
MODULE bar
 VAR
   q : boolean;
   p : boolean;

MODULE foo(c)
 DEFINE
   flag := c.q | c.p;
\end{nusmvCode}
%
Here, the value of `\code{m.flag}' is the logical \operator{OR} of
`\code{a.p}' and `\code{a.q}'.


Individual elements of an array are accessed in the typical fashion
with the index given in square brackets. See \ref{Index Subscript
Operator} for more information.


It is possible to refer to the name that the current module has been
instantiated to by using the \reserved{self} built-in
identifier.\index{self}
%
\begin{nusmvCode}
MODULE container(init_value1, init_value2)
  VAR c1 : counter(init_value1, self);
  VAR c2 : counter(init_value2, self);

MODULE counter(init_value, my_container)
  VAR v: 1..100;
  ASSIGN 
     init(v) := init_value;
  DEFINE 
     greatestCounterInContainer := v >= my_container.c1.v &
                                   v >= my_container.c2.v;

MODULE main
  VAR c : container(14, 7);
  SPEC
    c.c1.greatestCounterInContainer;
\end{nusmvCode}
%
In this example an instance of the module \code{container} is passed
to the sub-module \code{counter}. 
%
In the \code{main} module, \code{c} is declared to be an instance of
the module \code{container}, which declares two instances of the module
\code{counter}.
%
Every instance of the \code{counter} module has a define
\code{greatestCounterInContainer} which specifies the condition when
this particular \code{counter} has the greatest value in the container
it belongs to.  
%
So a \code{counter} needs access to the parent \code{container} to
access all the \code{counters} in the \code{container}.


%-----    ------
\subsection{Processes}
\label{Processes}
\index{process keyword}
\index{processes}
%
\fbox{\parbox[t]{\textwidth} {
\emph{Important!} \\
Since \nusmv version 2.5.0 processes are \emph{deprecated}.
In future versions of \nusmv processes may be no longer supported,
and only synchronous FSM will be supported by the input language.
Modeling of asynchronous processes will have to be resolved at
higher level.
}}
%
%
\vspace{3mm}

Processes are used to model interleaving
concurrency. A \dfn{process} is a module which is instantiated
using the keyword `\reserved{process}' (see \sref{MODULE
Instantiations}).
%
The program executes a step by non-deterministically choosing a
process, then executing all of the assignment statements in that
process in parallel. 
%
It is implicit that if a given variable is not assigned by the
process, then its value remains unchanged.
%
Note that only assignments of the form
%
\begin{nusmvCode}
ASSIGN next(\emph{var_name}) := \ldots ;
\end{nusmvCode}
%
are influenced by processes. All other kinds of assignments and
all constraints (such as \code{TRANS}, \code{INVAR}, etc) are always in
force, independent of which process is selected for execution.

Each instance of a process has a special \Boolean variable associated
with it, called \code{running}.\index{\code{running}}
%
The value of this variable is \code{TRUE} if and only if the process
instance is currently selected for execution.
%
No two processes may be running at the same time.

Note that (only) in the presence of processes NuSMV internally
declares special variables \code{running}
and \code{\_process\_selector\_}. These names should NOT be used in
user's own declarations (when processes are used), but they can be
referenced for example in the transition relation of a module.

Furthermore, if the user declares \code{N} processes, there will be
\code{N+1} processes allocated, as the module \code{main} has always its 
own process associated. In the following example there are three
process, \code{p1}, \code{p2} and \code{main}:
%
\begin{nusmvCode}
MODULE my_module
  -- my module definition...

MODULE main
  VAR 
    p1 : process my_module;
    p2 : process my_module;

\end{nusmvCode}
%

%-----    ------
\subsection{A Program and the \code{main} Module}
\label{Main Module}
\index{main module}
%
\index{syntax rules!main program}
The syntax of a \nusmv program is:
%
\begin{Grammar}
program :: module_list

module_list ::
          module
        | module_list module
\end{Grammar}
%
There must be one module with the name \code{main} and no formal
parameters.
%
The module \code{main} is the one evaluated by the interpreter.

%-----    ------
\subsection{Namespaces and Constraints on Declarations}
\label{Namespaces}
\index{namespaces}
\index{declarations}
%
Identifiers in the \nusmv input language may reference five different
entities: modules, variables, defines, module instances, and symbolic
constants.

Module identifiers have their own separate namespace. 
%
Module identifiers can be used in \grammar{module type specifiers}
only, and no other kind of identifiers can be used there (see
\sref{MODULE Instantiations}).
%
Thus, module identifiers may be equal to other kinds of identifiers
without making the program ambiguous. 
%
However, no two modules should be declared with the same
identifier. 
%
Modules cannot be declared in other modules, therefore they are always
referenced by simple \grammar{identifiers}.

Variable, define, and module instance identifiers are introduced in a
program when the module containing their declarations is
instantiated. 
%
Inside this module the variables, defines and module instances may be
referenced by the simple \grammar{identifiers}.
%
Inside other modules, their simple identifiers should be preceded by
the identifier of the module instance containing their declaration and
\code{\textbf{.}} (\code{<DOT>}).
%
Such identifiers are called \grammar{complex identifier}.
%
The \emph{full identifier} is a \grammar{complex identifier} which
references a variable, define, or a module instance from inside the
\code{main} module. 

Let us consider the following:
%
\begin{nusmvCode}
MODULE main
  VAR a : boolean;
  VAR b : foo;
  VAR c : moo;

MODULE foo
  VAR q : boolean;
      e : moo;

MODULE moo
  DEFINE f := 0 < 1;

MODULE not_used
  VAR n : boolean;
  VAR t : used;

MODULE used
  VAR k : boolean;
\end{nusmvCode}
%
The full identifier of the variable \code{a} is \code{a}, the full
identifier of the variable \code{q} (from the module \code{foo}) is
\code{b.q}, the full identifier of the module instance \code{e} (from
the module \code{foo}) is \code{b.e}, the full identifiers of the
define \code{f} (from the module \code{moo}) are \code{b.e.f} and
\code{c.f}, because two module instances contain this define.
%
Notice that, the variables \code{n} and \code{k} as well as the module
instance \code{t} do not have full identifiers because they cannot be
accessed from \code{main} (since the module \code{not\_used} is not
instantiated).

In the \nusmv language, variable, define, and module instances belong
to one namespace, and no two full identifiers of different variable,
define, or module instances should be equal. Also, none of them can be
redefined.

A \grammar{symbolic constant} can be introduced by a variable
declaration if its type specifier enumerates the \grammar{symbolic
constant}. For example, the variable declaration
%
\begin{nusmvCode}
  VAR a : \{OK, FAIL, waiting\};
\end{nusmvCode}
%
declares the variable \code{a} as well as the \grammar{symbolic
constants} \code{OK}, \code{FAIL} and \code{waiting}. 
%
The full identifiers of the \grammar{symbolic constants} are equal to
their simple \grammar{identifiers} with the additional condition --
the variable whose declaration declares the \grammar{symbolic
constants} also has a full identifier.

\grammar{Symbolic constants} have a separate namespace, so their
identifiers may potentially be equal, for example, variable
identifiers. 
%
It is an error, if the same identifier in an expression can
simultaneously refer to a \grammar{symbolic constant} and a variable
or a define.
%
A \grammar{symbolic constant} may be declared an arbitrary number of
times, but it must be declared at least once, if it is used in an
expression.

%-----    ------
\subsection{Context}
\label{Context}
\index{context}

Every module instance has its own \emph{context}, in which all
expressions are analyzed. 
%
The context can be defined as the full identifiers of variables
declared in the module without their simple identifiers.
%
Let us consider the following example:
%
\begin{nusmvCode}
MODULE main
  VAR a : foo;
  VAR b : moo;

MODULE foo
  VAR c : moo;

MODULE moo
  VAR d : boolean;
\end{nusmvCode}
%
The context of the module \code{main} is \code{`'}(empty)\footnote{
The module \code{main} is instantiated with the so called empty
identifier which cannot be referenced in a program.}, the context of
the module instance \code{a} (and inside the module \code{foo}) is
\code{`a.'}, the contexts of module \code{moo} may be \code{`b.'}  (if
the module instance \code{b} is analyzed) and \code{`a.c.'} (if the
module instance \code{a.c} is analyzed).

%-----    ------
\subsection{\code{ISA} Declarations}
\label{ISA Declarations}
\index{ISA declarations}
%
There are cases in which some parts of a module could be shared among
different modules, or could be used as a module themselves. 
%
In \nusmv it is possible to declare the common parts as separate
modules, and then use the \reserved{ISA} declaration to import the
common parts inside a module declaration. 
%
The syntax of an \grammar{isa\_declaration} is as follows:
%
\begin{Grammar}
isa_declaration :: \reserved{ISA} identifier
\end{Grammar}
%
where \grammar{identifier} must be the name of a declared module. 
%
The \grammar{ISA\_declaration} can be thought as a simple macro
expansion command, because the body of the module referenced by an
\code{ISA} command is replaced to the \grammar{ISA\_declaration}.

\textbf{Warning:} \reserved{ISA} is a deprecated feature and will be
removed from future versions of \nusmv. Therefore, avoid the use of
\grammar{ISA\_declarations}. Use module instances instead.

%-------------------- SPECIFICATIONS -----------------------------------

\section{Specifications}
%
The specifications to be checked on the FSM can be expressed in
temporal logics like Computation Tree Logic CTL, Linear Temporal Logic
LTL extended with Past Operators, and Property Specification Language
(PSL) \cite{PSLLRM} that includes CTL and LTL with Sequencial Extended
Regular Expressions (SERE), a variant of classical regular expressions.
%
It is also possible to analyze quantitative characteristics of the FSM
by specifying real-time CTL specifications. 
%
Specifications can be positioned within modules, in which case they
are preprocessed to rename the variables according to their context.

CTL and LTL specifications are evaluated by \nusmv in order to
determine their truth or falsity in the FSM. 
%
When a specification is discovered to be false, \nusmv constructs and
prints a counterexample, i.e. a trace of the FSM that falsifies the
property.

%-----    ------
\subsection{CTL Specifications}
\label{CTL Specifications}
\index{ CTL specifications}
%
A CTL specification is given as a formula in the temporal logic CTL,
introduced by the keyword `\reserved{CTLSPEC}' (however, deprecated
keyword `\reserved{SPEC}' can be used instead.)
%
The syntax of this specification is:
%
\begin{Grammar}
ctl_specification :: \reserved{CTLSPEC} ctl_expr [\textbf{;}]
                   | \reserved{SPEC} ctl_expr [\textbf{;}]
                   | \reserved{CTLSPEC} \reserved{NAME} name \textbf{:=} ctl_expr [\textbf{;}]
                   | \reserved{SPEC} \reserved{NAME} name \textbf{:=} ctl_expr [\textbf{;}]

\end{Grammar}
%
The syntax of CTL formulas recognized by \nusmv is as follows:
%
\begin{Grammar}
ctl_expr ::
    simple_expr                 -- a simple boolean expression
    | \textbf{(} ctl_expr \textbf{)}
    | \operator{!} ctl_expr                -- logical not
    | ctl_expr \operator{\&} ctl_expr       -- logical and
    | ctl_expr \operator{|} ctl_expr       -- logical or
    | ctl_expr \operator{xor} ctl_expr     -- logical exclusive or
    | ctl_expr \operator{xnor} ctl_expr    -- logical NOT exclusive or
    | ctl_expr \operator{->} ctl_expr      -- logical implies
    | ctl_expr \operator{<->} ctl_expr     -- logical equivalence
    | \reserved{EG} ctl_expr               -- exists globally
    | \reserved{EX} ctl_expr               -- exists next state
    | \reserved{EF} ctl_expr               -- exists finally
    | \reserved{AG} ctl_expr               -- forall globally
    | \reserved{AX} ctl_expr               -- forall next state
    | \reserved{AF} ctl_expr               -- forall finally
    | \reserved{E} \textbf{[} ctl_expr \reserved{U} ctl_expr \textbf{]} -- exists until
    | \reserved{A} \textbf{[} ctl_expr \reserved{U} ctl_expr \textbf{]} -- forall until
\end{Grammar}
%
Since \grammar{simple\_expr} cannot contain the \operator{next}
operator, \grammar{ctl\_expr} cannot contain it either.
%
The \grammar{ctl\_expr} should also be a \Boolean expression.

Intuitively the semantics of CTL operators is as follows:
% 
\begin{itemize}
  \item \code{\reserved{EX} \textit{p}}
        is true in a state $s$ if \emph{there exists} a state
        $s^\prime$ such that a transition goes from $s$ to $s^\prime$
        and \textit{p} is true in $s^\prime$.
  \item \code{\reserved{AX} \textit{p}}
        is true in a state $s$ if \emph{for all} states $s^\prime$
        where there is a transition from $s$ to $s^\prime$, \textit{p}
        is true in $s^\prime$.
  \item \code{\reserved{EF} \textit{p}}
       is true in a state $s_0$ if \emph{there exists} a series of
       transitions $s_0 \rightarrow s_1$, $s_1 \rightarrow s_2$,
       \ldots, $s_{n-1} \rightarrow s_n$ such that \textit{p} is true
       in $s_n$.
  \item \code{\reserved{AF} \textit{p}}
       is true in a state $s_0$ if \emph{for all} series of
       transitions $s_0 \rightarrow s_1$, $s_1 \rightarrow s_2$,
       \ldots, $s_{n-1} \rightarrow s_n$ \textit{p} is true in $s_n$.

  \item \code{\reserved{EG} \textit{p}}
       is true in a state $s_0$ if \emph{there exists} an infinite
       series of transitions $s_0 \rightarrow s_1$, $s_1 \rightarrow
       s_2$, \ldots~such that \textit{p} is true in \emph{every}
       $s_i$.
  \item \code{\reserved{AG} \textit{p}}
       is true in a state $s_0$ if \emph{for all} infinite series of
       transitions $s_0 \rightarrow s_1$, $s_1 \rightarrow s_2$,
       \ldots~\textit{p} is true in \emph{every} $s_i$.
  \item \code{\textbf{E[}\textit{p} \textbf{U} \textit{q}\textbf{]}}
       is true in a state $s_0$ if \emph{there exists} a series of
       transitions $s_0 \rightarrow s_1$, $s_1 \rightarrow s_2$,
       \ldots, $s_{n-1} \rightarrow s_n$ such that \textit{p} is true
       in \emph{every} state from $s_0$ to $s_{n-1}$ and \textit{q} is
       true in state $s_n$.
  \item \code{\textbf{A[}\textit{p} \textbf{U} \textit{q}\textbf{]}}
       is true in a state $s_0$ if \emph{for all} series of
       transitions $s_0 \rightarrow s_1$, $s_1 \rightarrow s_2$,
       \ldots, $s_{n-1} \rightarrow s_n$ \textit{p} is true in
       \emph{every} state from $s_0$ to $s_{n-1}$ and \textit{q} is
       true in state $s_n$.
\end{itemize}
A CTL formula is true if it is true in \emph{all} initial states.

For a detailed description about the semantics of \emph{PSL}
operators, please see \cite{PSLLRM}. 

%-----    ------
\subsection{Invariant Specifications}
\label{INVAR Specifications}
\index{INVARSPEC Specifications}
\index{Invariant Specifications}

It is also possible to specify invariant specifications with special
constructs. Invariants are propositional formulas which must hold
invariantly in the model.
%
The corresponding command is \reserved{INVARSPEC}, with syntax:
%
\begin{Grammar}
invar_specification :: \reserved{INVARSPEC} next_expr \textbf{;}
                       \reserved{INVARSPEC} \reserved{NAME} name \textbf{:=} next_expr [\textbf{;}]
\end{Grammar}
%
This statement is intuitively equivalent to 
%
\begin{Grammar}
SPEC  AG simple_expr ;
\end{Grammar}
%
but can be checked by a specialised algorithm during reachability
analysis and Invariant Specifications can contain \operator{next}
operators.  Fairness constraints are not taken into account during
invariant checking.

%-----    ------
\subsection{LTL Specifications}
\label{LTL Specifications}
\index{LTL Specifications}
%
LTL specifications are introduced by the keyword
\reserved{LTLSPEC}. 
%
The syntax of this specification is:
%
\begin{Grammar}
ltl_specification :: \reserved{LTLSPEC} ltl_expr [\textbf{;}]
                     \reserved{LTLSPEC} \reserved{NAME} name \textbf{:=} ltl_expr [\textbf{;}]
\end{Grammar}
%
The syntax of LTL formulas recognized by \nusmv is as follows:
\begin{Grammar}
ltl_expr ::
    simple_expr              -- a simple boolean expression
    | \textbf{(} ltl_expr \textbf{)}
    | \operator{!} ltl_expr             -- logical not
    | ltl_expr \operator{\&} ltl_expr    -- logical and
    | ltl_expr \operator{|} ltl_expr    -- logical or
    | ltl_expr \operator{xor} ltl_expr  -- logical exclusive or
    | ltl_expr \operator{xnor} ltl_expr -- logical NOT exclusive or
    | ltl_expr \operator{->} ltl_expr   -- logical implies
    | ltl_expr \operator{<->} ltl_expr  -- logical equivalence
    -- FUTURE
    | \reserved{X} ltl_expr             -- next state
    | \reserved{G} ltl_expr             -- globally
    | \reserved{F} ltl_expr             -- finally
    | ltl_expr \reserved{U} ltl_expr    -- until
    | ltl_expr \reserved{V} ltl_expr    -- releases
    -- PAST
    | \reserved{Y} ltl_expr             -- previous state
    | \reserved{Z} ltl_expr             -- not previous state not
    | \reserved{H} ltl_expr             -- historically
    | \reserved{O} ltl_expr             -- once 
    | ltl_expr \reserved{S} ltl_expr    -- since
    | ltl_expr \reserved{T} ltl_expr    -- triggered
\end{Grammar}
%
Intuitively the semantics of LTL operators is as follows:
% 
\begin{itemize}
  \item \code{\reserved{X} \textit{p}}
        is true at time $t$ if \textit{p} is true at time $t+1$.
  \item \code{\reserved{F} \textit{p}}
        is true at time $t$ if \textit{p} is true at \emph{some} time
        $t^\prime \ge t$.
  \item \code{\reserved{G} \textit{p}}
        is true at time $t$ if \textit{p} is true at \emph{all} times
        $t^\prime \ge t$.
  \item \code{\textit{p} \reserved{U} \textit{q}}
        is true at time $t$ if \textit{q} is true at \emph{some} time
        $t^\prime \ge t$, and \emph{for all} time $t^{\prime\prime}$
        (such that $t \le t^{\prime\prime} < t^\prime$) \textit{p} is
        true.
  \item \code{\textit{p} \reserved{V} \textit{q}}
        is true at time $t$ if \textit{q} holds at \emph{all} time steps $t^\prime
        \geq t$ up to and including the time step $t^{\prime\prime}$
        where \textit{p} also holds. Alternatively, it may be the case that \textit{p}
        \emph{never} holds in which case \textit{q} must hold in \emph{all} time
        steps $t^\prime \geq t$.
  \item \code{\reserved{Y} \textit{p}} is true at time $t > t_0$ if \textit{p} holds
        at time $t - 1$. \code{\reserved{Y} \textit{p}} is \emph{false} at time $t_0$.
  \item \code{\reserved{Z} \textit{p}} is equivalent to \code{\reserved{Y} \textit{p}}
        with the exception that the expression is \emph{true} at time $t_0$.
  \item \code{\reserved{H} \textit{p}} is true at time $t$ if \textit{p} holds in \emph{all}
        previous time steps $t^\prime \leq t$.
  \item \code{\reserved{O} \textit{p}} is true at time $t$ if \textit{p} held in \emph{at least one}
        of the previous time steps $t^\prime \leq t$.
  \item \code{\textit{p} \reserved{S} \textit{q}} is true at time $t$ if \textit{q} held at time
        $t^\prime \leq t$ and \textit{p} holds in \emph{all} time steps $t^{\prime\prime}$ such
        that $t^\prime < t^{\prime\prime} \le t$.
  \item \code{\textit{p} \reserved{T} \textit{q}} is true at time $t$ if \textit{p} held at time
        $t^\prime \leq t$ and \textit{q} holds in \emph{all} time steps $t^{\prime\prime}$ such
        that $t^\prime \le t^{\prime\prime} \le t$. Alternatively, if \textit{p} has \emph{never}
        been true, then \textit{q} must hold in all time steps $t^{\prime\prime}$ such
        that $t_0 \le t^{\prime\prime} \le t$
\end{itemize}
An LTL formula is true if it is true at the initial time $t_0$.

In \nusmv, LTL specifications can be analyzed both by means of
BDD-based reasoning, or by means of SAT-based bounded model
checking. 
%
In the case of BDD-based reasoning, \nusmv proceeds according to
\cite{CGH97}. 
%
For each LTL specification, a tableau of the behaviors falsifying the
property is constructed, and then synchronously composed with the
model. 
%
With respect to \cite{CGH97}, the approach is fully integrated within
\nusmv, and allows full treatment of past temporal operators. 
%
Note that the counterexample is generated in such a way to show that the falsity
of a LTL specification may contain state variables which have been
introduced by the tableau construction procedure.

In the case of SAT-based reasoning, a similar tableau construction is
carried out to encode the paths of limited length, violating the
property.
%
\nusmv generates a propositional satisfiability problem, that is then
tackled by means of an efficient SAT solver \cite{BCCZ99}.

In both cases, the tableau constructions are completely transparent to
the user.

%-----    ------
\subsubsection{Important Difference Between BDD and SAT Based LTL Model Checking}
\label{Important Difference Between BDD and SAT Based LTL Model Checking}
\index{Important Difference Between BDD and SAT Based LTL Model Checking}

If a FSM to be checked it not total (i.e. has deadlock state) the
model checking may return different results for the same LTL
specification depending on the verification engine used.  For
example, for below model:
%
\begin{nusmvCode}
MODULE main
VAR s : boolean;
TRANS s = TRUE
LTLSPEC G (s = TRUE)
\end{nusmvCode}
%
the LTL specification is proved valid by BDD-based model checking but
is violated by SAT-based bounded model checking. The counter-example
found consists of one state \code{s=FALSE}.

This difference between the results is caused by the fact that BDD
model checking investigates only \emph{infinite} paths whereas
SAT-based model checking is able to deal also with
\emph{finite} paths.
%
Apparently infinite paths cannot ever have \code{s=FALSE} as then the
transition relation will not hold between the consecutive states in
the path.
%
A \emph{finite} path consisting of just one state \code{s=FALSE}
violates the specification \code{G (s = TRUE)} and is still consistent
with the FSM as the transition relation is not taken ever and there is
not initial condition to violate. Note however that this state is a
deadlock and cannot have consecutive states.

In order to make SAT-based bound model checking ignore finite paths
it is enough to add a fairness condition to the \code{main} module:
%
\begin{nusmvCode}
JUSTICE TRUE;
\end{nusmvCode}
%
Being limited to fair paths, SAT-based bounded model checking cannot
find a finite counter-example and results of model checking become
consistent with BDD-based model checking.

%-----    ------
\subsection{Real Time CTL Specifications and Computations}
\label{Real Time CTL Specifications and Computations}
\index{Real Time CTL Specifications and Computations}

\nusmv allows for Real Time CTL specifications~\cite{EMSS91}.
%
\nusmv assumes that each transition takes unit time for
execution.
%
RTCTL extends the syntax of CTL path expressions with the following
bounded modalities:
%
\begin{Grammar}
rtctl_expr ::
        ctl_expr
      | \reserved{EBF} range rtctl_expr
      | \reserved{ABF} range rtctl_expr
      | \reserved{EBG} range rtctl_expr
      | \reserved{ABG} range rtctl_expr
      | \reserved{A} \textbf{[} rtctl_expr \reserved{BU} range rtctl_expr \textbf{]}
      | \reserved{E} \textbf{[} rtctl_expr \reserved{BU} range rtctl_expr \textbf{]}
range  :: integer_number \textbf{..} integer_number
\end{Grammar}
%
Given ranges must be non-negative.\\
%
Intuitively, the semantics of the RTCTL operators is as follows:\\
%
\begin{itemize}
  \item \code{\reserved{EBF} \textit{m}\textbf{..}\textit{n} \textit{p}}
        requires that there exists a path starting from a state, such
        that property \textit{p} holds in a future time instant
        \textit{i}, with $m \leq i \leq n$
  \item \code{\reserved{ABF} \textit{m}\textbf{..}\textit{n} \textit{p}}
        requires that for all paths starting from a state, property
        \textit{p} holds in a future time instant \textit{i}, with $m
        \leq i \leq n$
  \item \code{\reserved{EBG} \textit{m}\textbf{..}\textit{n} \textit{p}}
        requires that there exists a path starting from a state, such
        that property \textit{p} holds in all future time instants
        \textit{i}, with $m \leq i \leq n$
  \item \code{\reserved{ABG} \textit{m}\textbf{..}\textit{n} \textit{p}}
        requires that for all paths starting from a state, property
        \textit{p} holds in all future time instants \textit{i}, with
        $m \leq i \leq n$
  \item \code{\reserved{E} \textbf{[} \textit{p} \reserved{BU} \textit{m}\textbf{..}\textit{n} \textit{q} \textbf{]}}
        requires that there exists a path starting from a state, such
        that property \textit{q} holds in a future time instant
        \textit{i}, with $m \leq i \leq n$, and property \textit{p}
        holds in all future time instants \textit{j}, with $m \leq j <
        i$
  \item \code{\reserved{A} \textbf{[} \textit{p} \reserved{BU} \textit{m}\textbf{..}\textit{n} \textit{q} \textbf{]}},
        requires that for all paths starting from a state, property
        \textit{q} holds in a future time instant \textit{i}, with $m
        \leq i \leq n$, and property \textit{p} holds in all future
        time instants \textit{j}, with $m \leq j < i$
\end{itemize}
%
Real time CTL specifications can be defined with the following syntax,
which extends the syntax for CTL specifications. (keyword
`\reserved{SPEC}' is deprecated)
%
\begin{Grammar}
rtctl_specification :: \reserved{CTLSPEC} rtctl_expr [\textbf{;}]
                     | \reserved{SPEC} rtctl_expr [\textbf{;}]
                     | \reserved{CTLSPEC} \reserved{NAME} name \textbf{:=} rtctl_expr [\textbf{;}]
                     | \reserved{SPEC} \reserved{NAME} name \textbf{:=} rtctl_expr [\textbf{;}]
\end{Grammar}
%
With the \reserved{COMPUTE} statement, it is also possible to compute
quantitative information on the FSM. 
%
In particular, it is possible to compute the exact bound on the delay
between two specified events, expressed as CTL formulas. 
%
The syntax is the following:
%
\begin{Grammar}
compute_specification :: \reserved{COMPUTE} compute_expr [\textbf{;}]
                         \reserved{COMPUTE} \reserved{NAME} name \textbf{:=} compute_expr [\textbf{;}]
\end{Grammar}
%
where
%
\begin{Grammar}
compute_expr :: \reserved{MIN} \textbf{[} rtctl_expr \textbf{,} rtctl_expr \textbf{]}
              | \reserved{MAX} \textbf{[} rtctl_expr \textbf{,} rtctl_expr \textbf{]}
\end{Grammar}
%
\code{\reserved{MIN} \textbf{[}\textit{start} \textbf{,}
\textit{final}]} returns the length of the shortest path from a state
in \textit{start} to a state in \textit{final}.
%
For this, the set of states reachable from \textit{start} is
computed. 
%
If at any point, we encounter a state satisfying \textit{final}, we
return the number of steps taken to reach the state. 
%
If a fixed point is reached and no computed states intersect
\textit{final} then \dfn{infinity} is returned.

\noindent\code{\reserved{MAX} \textbf{[}\textit{start} \textbf{,}
\textit{final}]} returns the length of the longest path from a state
in \textit{start} to a state in \textit{final}. 
%
If there exists an infinite path beginning in a state in
\textit{start} that never reaches a state in \textit{final}, then
\textit{infinity} is returned. If any of the initial or final states 
is empty, then \textit{undefined} is returned.

It is important to remark here that if the FSM is not total (i.e. it
contains deadlock states) \reserved{COMPUTE} may produce wrong
results. It is possible to check the FSM against deadlock states by
calling the command \command{check\_fsm}.


%-----    ------
\subsection{PSL Specifications}
\label{PSL Specifications}
\index{PSL Specifications}
%
\nusmv allows for PSL specifications as from version 1.01 of PSL
Language Reference Manual \cite{PSLLRM}. PSL specifications
are introduced by the keyword ``\code{PSLSPEC}''. The syntax of this
declaration (as from the PSL parsers distributed by IBM, \cite{PSLparser}) is:
%
\begin{Grammar}
pslspec_declaration :: \reserved{PSLSPEC} psl_expr [\textbf{;}]
                       \reserved{PSLSPEC} \reserved{NAME} name \textbf{:=} psl_expr [\textbf{;}]
\end{Grammar}
%
where
%
\begin{Grammar}
psl_expr ::
   psl_primary_expr
 | psl_unary_expr
 | psl_binary_expr
 | psl_conditional_expr
 | psl_case_expr
 | psl_property
\end{Grammar}
%
The first five classes define the building blocks for
\code{psl\_property} and provide means of combining
instances of that class; they are defined as follows:

\begin{Grammar}
psl_primary_expr ::
   number                              ;; a numeric constant
 | boolean                             ;; a boolean constant
 | word                                ;; a word constant
 | var_id                              ;; a variable identifier
 | \textbf{\{} psl_expr \textbf{,} ... \textbf{,} psl_expr \textbf{\}}
 | \textbf{\{} psl_expr "\{" psl_expr \textbf{,} ... \textbf{,} "psl_expr" \textbf{\}}\textbf{\}}
 | \textbf{(} psl_expr \textbf{)}

psl_unary_expr ::
   \operator{+} psl_primary_expr     
 | \operator{-} psl_primary_expr  
 | \operator{!} psl_primary_expr  
 | \operator{bool} \textbf{(} psl_expr \textbf{)}
 | \operator{word1} \textbf{(} psl_expr \textbf{)}
 | \operator{uwconst} \textbf{(} psl_expr, psl_expr \textbf{)}
 | \operator{swconst} \textbf{(} psl_expr, psl_expr \textbf{)}
 | \operator{sizeof} \textbf{(} psl_expr \textbf{)}
 | \operator{toint} \textbf{(} psl_expr \textbf{)}
 | \operator{signed} \textbf{(} psl_expr \textbf{)}
 | \operator{unsigned} \textbf{(} psl_expr \textbf{)}
 | \operator{extend} \textbf{(} psl_expr, psl_primary_expr \textbf{)}
 | \operator{resize} \textbf{(} psl_expr, psl_primary_expr \textbf{)}
 | \operator{select} \textbf{(} psl_expr, psl_expr, psl_expr \textbf{)}

psl_binary_expr ::
   psl_expr \operator{+} psl_expr    
 | psl_expr \operator{union} psl_expr 
 | psl_expr \operator{in} psl_expr 
 | psl_expr \operator{-} psl_expr   
 | psl_expr \operator{*}psl_expr   
 | psl_expr \operator{/} psl_expr   
 | psl_expr \operator{\%} psl_expr 
 | psl_expr \operator{==} psl_expr    
 | psl_expr \operator{!=} psl_expr  
 | psl_expr \operator{<} psl_expr       
 | psl_expr \operator{<=} psl_expr       
 | psl_expr \operator{>} psl_expr       
 | psl_expr \operator{>=} psl_expr       
 | psl_expr \operator{&} psl_expr 
 | psl_expr \operator{|} psl_expr 
 | psl_expr \operator{xor} psl_expr 
 | psl_expr \operator{xnor} psl_expr 
 | psl_expr \operator{<<} psl_expr 
 | psl_expr \operator{>>} psl_expr 
 | psl_expr \operator{::} psl_expr 
psl_conditional_expr ::
 psl_expr \textbf{?} psl_expr \textbf{:} psl_expr 
psl_case_expr ::
 \reserved{case}
     psl_expr \textbf{:} psl_expr \textbf{;}
     ...
     psl_expr \textbf{:} psl_expr \textbf{;}
 \reserved{endcase}
\end{Grammar}
%
Among the subclasses of \code{psl\_expr} we depict the class
\code{psl\_bexpr} that will be used in the following to identify purely
boolean, i.e. not temporal, expressions. The class of PSL properties
\code{psl\_property} is defined as follows:
%
\begin{Grammar}
psl_property :: 
   replicator psl_expr ;; a replicated property 
 | FL_property \reserved{abort} psl_bexpr
 | psl_expr \operator{<->} psl_expr
 | psl_expr \operator{->} psl_expr
 | FL_property       
 | OBE_property      
replicator :: 
   \reserved{forall} var_id [index_range] \reserved{in} value_set \textbf{:} 
index_range :: 
   \textbf{[} range \textbf{]} 
range :: 
   low_bound \textbf{:} high_bound 
low_bound :: 
   number              
 | identifier         
high_bound :: 
   number 
 | identifier
 | \reserved{inf}             ;; inifite high bound 
value_set :: 
   \textbf{\{} value_range \textbf{,} ... \textbf{,} value_range \textbf{\}}
 | \reserved{boolean}
value_range :: 
   psl_expr
 | range
\end{Grammar}
%
The instances of \code{FL\_property} are temporal properties built
using LTL operators and SEREs operators, and are defined as follows:
%

\begin{Grammar}
FL_property ::
 ;; PRIMITIVE LTL OPERATORS
   \reserved{X} FL_property                      
 | \reserved{X!} FL_property                     
 | \reserved{F} FL_property                      
 | \reserved{G} FL_property                      
 | \textbf{[} FL_property \reserved{U} FL_property \textbf{]}  
 | \textbf{[} FL_property \reserved{W} FL_property \textbf{]}  
 ;; SIMPLE TEMPORAL OPERATORS
 | \reserved{always} FL_property                 
 | \reserved{never} FL_property                  
 | \reserved{next} FL_property                   
 | \reserved{next!} FL_property                  
 | \reserved{eventually!} FL_property            
 | FL_property \reserved{until!} FL_property     
 | FL_property \reserved{until} FL_property      
 | FL_property \reserved{until!_} FL_property                                     
 | FL_property \reserved{until_} FL_property     
 | FL_property \reserved{before!} FL_property    
 | FL_property \reserved{before} FL_property     
 | FL_property \reserved{before!_} FL_property   
 | FL_property \reserved{before_} FL_property    
 ;; EXTENDED NEXT OPERATORS
 | \reserved{X} [number] \textbf{(} FL_property \textbf{)}
 | \reserved{X!} [number] \textbf{(} FL_property \textbf{)}                     
 | \reserved{next} [number] \textbf{(} FL_property \textbf{)}                   
 | \reserved{next!} [number] \textbf{(} FL_property \textbf{)}                  
 ;;
 | \reserved{next_a} [range] \textbf{(} FL_property \textbf{)}
 | \reserved{next_a!} [range] \textbf{(} FL_property \textbf{)}
 | \reserved{next_e} [range] \textbf{(} FL_property \textbf{)}
 | \reserved{next_e!} [range] \textbf{(} FL_property \textbf{)}
 ;;
 | \reserved{next_event!} \textbf{(} psl_bexpr \textbf{)} \textbf{(} FL_property \textbf{)}
 | \reserved{next_event} \textbf{(} psl_bexpr \textbf{)} \textbf{(} FL_property \textbf{)}
 | \reserved{next_event!} \textbf{(} psl_bexpr \textbf{)} \textbf{[} number \textbf{]}  \textbf{(} FL_property \textbf{)}
 | \reserved{next_event} \textbf{(} psl_bexpr \textbf{)} \textbf{[} number \textbf{]}  \textbf{(} FL_property \textbf{)}
 ;;
 | \reserved{next_event_a!} \textbf{(} psl_bexpr \textbf{)} \textbf{[}psl_expr\textbf{]}  \textbf{(} FL_property \textbf{)}
 | \reserved{next_event_a} \textbf{(} psl_bexpr \textbf{)} \textbf{[}psl_expr\textbf{]}  \textbf{(} FL_property \textbf{)}
 | \reserved{next_event_e!} \textbf{(} psl_bexpr \textbf{)} \textbf{[}psl_expr\textbf{]}  \textbf{(} FL_property \textbf{)}
 | \reserved{next_event_e} \textbf{(} psl_bexpr \textbf{)} \textbf{[}psl_expr\textbf{]}  \textbf{(} FL_property \textbf{)}
 ;; OPERATORS ON SEREs
 | sequence \textbf{(} FL_property \textbf{)}
 | sequence \textbf{|->} sequence [\textbf{!}]
 | sequence \textbf{|=>} sequence [\textbf{!}]
 ;;
 | \reserved{always} sequence
 | \reserved{G} sequence
 | \reserved{never} sequence
 | \reserved{eventually!} sequence
 ;;
 | \reserved{within!} \textbf{(} sequence_or_psl_bexpr \textbf{,} psl_bexpr \textbf{)} sequence
 | \reserved{within} \textbf{(} sequence_or_psl_bexpr \textbf{,} psl_bexpr \textbf{)} sequence
 | \reserved{within!_} \textbf{(} sequence_or_psl_bexpr \textbf{,} psl_bexpr \textbf{)} sequence
 | \reserved{within_} \textbf{(} sequence_or_psl_bexpr \textbf{,} psl_bexpr \textbf{)} sequence
 ;;
 | \reserved{whilenot!} \textbf{(} psl_bexpr \textbf{)} sequence
 | \reserved{whilenot} \textbf{(} psl_bexpr \textbf{)} sequence
 | \reserved{whilenot!_} \textbf{(} psl_bexpr \textbf{)} sequence
 | \reserved{whilenot_} \textbf{(} psl_bexpr \textbf{)} sequence
sequence_or_psl_bexpr ::
   sequence
 | psl_bexpr
\end{Grammar}
%
Please note that instances of \code{FL\_property} cannot be combined
with the ``\textbf{=}'', ``\textbf{!=}'' and ``\textbf{==}''.\\
%
Sequences, i.e. istances of class \code{sequence}, are defined as follows:
%
\begin{Grammar}
sequence ::
   \textbf{\{} SERE \textbf{\}}
SERE ::
   sequence
 | psl_bexpr
 ;; COMPOSITION OPERATORS
 | SERE \operator{;} SERE
 | SERE \operator{:} SERE
 | SERE \operator{&} SERE
 | SERE \operator{&&} SERE
 | SERE \operator{|} SERE
 ;; RegExp QUALIFIERS
 | SERE \textbf{[*} [count] \textbf{]}
 | \textbf{[*} [count] \textbf{]}
 | SERE \textbf{[+]}
 | \textbf{[+]}
 ;;
 | psl_bexpr \textbf{[=} count \textbf{]}
 | psl_bexpr \textbf{[->} count \textbf{]}
count ::
   number
 | range
\end{Grammar}
%
Istances of \code{OBE\_property} are CTL properties in the PSL style
and are defined as follows:
%
\begin{Grammar}
OBE_property ::
   \reserved{AX} OBE_property
 | \reserved{AG} OBE_property
 | \reserved{AF} OBE_property
 | \reserved{A} \textbf{[} OBE_property \reserved{U} OBE_property \textbf{]}
 | \reserved{EX} OBE_property
 | \reserved{EG} OBE_property
 | \reserved{EF} OBE_property
 | \reserved{E} \textbf{[} OBE_property \reserved{U} OBE_property \textbf{]}
\end{Grammar}
%
The \nusmv parser allows to input any specification based on the
grammar above, but currently, verification of PSL specifications is
supported only for the OBE subset, and for a subset of PSL for
which it is possible to define a translation into LTL. For the
specifications that belong to these subsets, it is possible to apply
all the verification techniques that can be applied to LTL and CTL
Specifications.


\section{Variable Order Input}
\label{Variable::Order::Input}
%
It is possible to specify the order in which variables should appear
in the BDD's generated by \nusmv. 
%
The file which gives the desired order can be read in using the
\commandopt{i} option in batch mode or by setting the
\envvar{input\_order\_file} environment variable in interactive mode.
%
 \footnote{Note that if the ordering is not provided by a user then
 \nusmv decides by itself how to order the variables. Two shell
 variables \varName{bdd\_static\_order\_heuristics} (see page
 \pageref{bdd_static_order_heuristics}) and
 \varName{vars\_order\_type} (see page \pageref{vars_order_type}) allow
 to control the ordering creation.}

%-----    ------
\subsection{Input File Syntax}
\label{Input File Syntax}
\index{ Input File Syntax}
%
The syntax for input files describing the desired variable ordering is
as follows, where the file can be considered as a list of variable
names, each of which must be on a separate line:
%
\begin{Grammar}
vars_list :: \textit{EMPTY}
         | var_list_item vars_list

var_list_item :: complex_identifier
             | complex_identifier \textbf{.} integer_number
\end{Grammar}
%
Where \textit{\texttt{EMPTY}} means parsing nothing.

This grammar allows for parsing a list of variable names of the
following forms:
%
\begin{nusmvCode}
Complete_Var_Name        -- to specify an ordinary variable
Complete_Var_Name[index] -- to specify an array variable element
Complete_Var_Name.NUMBER -- to specify a specific bit of a
                         -- scalar variable
\end{nusmvCode}
%
where \grammar{Complete\_Var\_Name} is just the name of the variable if
it appears in the module \code{MAIN}, otherwise it has the module
name(s) prepended to the start, for example:\\

\code{mod1.mod2...modN.varname}\\

\noindent where \code{varname} is a variable in \code{modN}, and
\code{modN.varname} is a variable in \code{modN-1}, and so
on. Note that the module name \code{main} is implicitely prepended
to every variable name and therefore must not be included in their
declarations.

\noindent Any variable which appears in the model file, but not the
ordering file is placed after all the others in the
ordering. Variables which appear in the ordering file but not the
model file are ignored. In both cases \nusmv displays a warning
message stating these actions.

Comments can be included by using the same syntax as regular \nusmv
files. That is, by starting the line with \texttt{--}.


%-----    ------
\subsection{Scalar Variables}
\label{Scalar Variables}
\index{Scalar Variables}
%
A variable, which has a finite range of values that it can take, is
encoded as a set of \Boolean variables (i.e.\@ bits). These boolean variables
represent the binary equivalents of all the possible values for the
scalar variable. Thus, a scalar variable that can take values from 0
to 7 would require three \Boolean variables to represent it.

It is possible not only to declare the position of a scalar variable
in the ordering file, but each of the \Boolean variables which
represent it.

\noindent If only the scalar variable itself is named then all the
boolean variables which are actually used to encode it are grouped
together in the BDD package.

\noindent Variables which are grouped together will always remain next
to each other in the BDD package and in the same order. 
%
When dynamic variable re-ordering is carried out, the group of
variables are treated as one entity and moved as such.

\noindent If a scalar variable is omitted from the ordering file then
it will be added at the end of the variable order and the specific-bit
variables that represent it will be grouped together. 
%
However, if any specific-bit variables have been declared in the
ordering file (see below) then these will not be grouped with the
remaining ones.

\noindent It is also possible to specify the location of specific bit variables
anywhere in the ordering.
%
This is achieved by first specifying the scalar variable name in the
desired location, then simply specifying
\texttt{Complete\_Var\_Name.i} at the position where you want that bit
variable to appear:
%
\begin{Grammar}
...
Complete\_Var\_Name
...
Complete\_Var\_Name.i
...
\end{Grammar}
%
The result of doing this is that the variable representing the
\textit{$i^{th}$} bit is located in a different position to the
remainder of the variables representing the rest of the bits. 
%
The specific-bit variables \textit{varname.0, ..., varname.i-1,
varname.i+1, ..., varname.N} are grouped together as before.

If any one bit occurs before the variable it belongs to, the remaining
specific-bit variables are not grouped together:
%
\begin{Grammar}
...
Complete\_Var\_Name.i
...
Complete\_Var\_Name
...
\end{Grammar}
%
The variable representing the \textit{$i^{th}$} bit is located at the
position given in the variable ordering and the remainder are located
where the scalar variable name is declared. In this case, the
remaining bit variables will not be grouped together.

\noindent This is just a short-hand way of writing each individual
specific-bit variable in the ordering file. The following are
equivalent:

\begin{small} 
\begin{tabular}{ll} \texttt{\ldots} &
\texttt{\ldots}\\ \texttt{Complete\_Var\_Name.0} &
\texttt{Complete\_Var\_Name.0}\\ \texttt{Complete\_Var\_Name.1} &
\texttt{Complete\_Var\_Name}\\ \texttt{$\vdots$} & \raisebox{1ex}{\texttt{\ldots}}\\
% Have to use \raisebox{} since \cdots display differently (not affected by \texttt{})
\texttt{Complete\_Var\_Name.N-1}\\ \texttt{\ldots} &
\end{tabular}
\end{small}

\noindent where the scalar variable \grammar{Complete\_Var\_Name}
requires N boolean variables to encode all the possible values that it
may take. 
%
It is still possible to then specify other specific-bit variables at
later points in the ordering file as before.\\

%-----    ------
\subsection{Array Variables}
\label{Array Variables}
\index{ Array Variables}

When declaring array variables in the ordering file, each individual
element must be specified separately.
%
It is not permitted to specify just the name of the array.
%
The reason for this is that the actual definition of an array in the
model file is essentially a shorthand method of defining a list of
variables that all have the same type. 
%
Nothing is gained by declaring it as an array over declaring each of
the elements individually, and there is no difference in terms of the
internal representation of the variables.


\section{Clusters Ordering}
%
When \nusmv builds a clusterized BDD-based FSM during model
construction, an initial simple clusters list is roughly constructed
by iterating through a \emph{list of variables}, and by constructing
the clusters by picking the transition relation associated to each
variable in the list. Later, the clusters list will be refined and
improved by applying the clustering alghorithm that the user
previoulsy selected (see partitioning methods at page
\ref{build::model} for further information).


In \cite{fm06}, Wendy Johnston and others from University of
Queensland, showed that choosing a good ordering for the initial list
of variables that is used to build the clusters list may lead to a
dramatic improvement of performances. They did experiments in a
modified version of \nusmv, by allowing the user to specify a variable
ordering to be used when constructing the initial clusters list. The
prototype code has been included in version 2.4.1, that offers the new
option \code{trans\_order\_file} to specify a file containing a
variable ordering (see at page \pageref{build::model} for further
information).


Grammar of the clusters ordering file is the same of variable ordering
file presented in section \ref{Variable::Order::Input} at page
\pageref{Variable::Order::Input}.

