% -*-latex-*-
\begin{nusmvCommand}{check\_property} {Checks a property into the current list of properties, 
  or a newly specified property}

\cmdLine{check\_property [-h] [-n number | -P "name"] | [(-c | -l | -i | -s | -q ) [-p "formula
   [IN context]"]]}

Checks the specified property taken from the property list, or adds
the new specified property and checks it. It is possible to check
\code{LTL, CTL, INVAR, PSL} and quantitative (\code{COMPUTE})
properties. Every newly inserted property is inserted and checked.

See variable \varName{use\_coi\_size\_sorting} for changing properties
verification order.

\begin{cmdOpt}
\opt{-n \parameter{\natnum{number}}}{Checks the property stored at the given index}

\opt{-P \parameter{\natnum{name}}}{Checks the property named \natnum{name} in
the property database.} 

\opt{-c}{Checks all the \code{CTL} properties not already checked. If
-p is used, the given formula is expected to be a \code{CTL}
formula.}

\opt{-l}{Checks all the \code{LTL} properties not already checked. If
-p is used, the given formula is expected to be a \code{LTL}
formula.}

\opt{-i}{Checks all the \code{INVAR} properties not already checked. If
-p is used, the given formula is expected to be a \code{INVAR}
formula.}

\opt{-s}{Checks all the \code{PSL} properties not already checked. If
\code{-p} is used, the given formula is expected to be a \code{PSL}
formula.}
 
\opt{-q}{Checks all the \code{COMPUTE} properties not already checked. If
-p is used, the given formula is expected to be a \code{COMPUTE}
formula.}

\opt{-p \parameter{"\anyexpr [IN context]"}}{Checks the formula
          specified on the command-line.  \code{context} is the module
          instance name which the variables in \anyexpr must be
          evaluated in.}

\end{cmdOpt}

\end{nusmvCommand}
