% -*-latex-*-
\begin{nusmvCommand} {check\_ltlspec\_bmc\_onepb} {Checks the given LTL specification, or all LTL specifications if no formula is given. Checking parameters are the single problem bound and the loopback value}

\cmdLine{check\_ltlspec\_bmc\_onepb [-h ] | [ -n idx | -p "formula"
    [IN context] | -P "name"] [-k length] [-l loopback] [-o filename]}

As command \command{check\_ltlspec\_bmc} but it produces only one
single problem with fixed bound and loopback values, with no iteration
of the problem bound from zero to max\_length.

\begin{cmdOpt}

\opt{-n \parameter{\natnum{\it index}}}{{\it index} is the numeric index of a valid LTL
specification formula actually located in the properties database.
The validity of {\it index} value is checked out by the system.}

\opt{-p \parameter{"\anyexpr [IN context]"}}{Checks the \anyexpr
specified on the command-line. \code{context} is the module instance
name which the variables in \anyexpr must be evaluated in.}

\opt{-P \parameter{\natnum{name}}}{Checks the LTL property named \natnum{name} in
the property database.} 

\opt{-k \parameter{\natnum{\it length}}}{{\it length} is the problem bound used when
generating the single problem. Only natural numbers are valid values
for this option. If no value is given the environment variable
\envvar{bmc\_length} is considered instead.}

\opt{-l \parameter{\set{\it loopback}{\range{0}{max\_length-1},
       \range{-1}{bmc\_length}, X, *}}}{The {\it loopback} value may be:}
       \tabItem{a natural number in (0, {\it max\_length-1}). A positive sign ('+') can 
       be also used as prefix of the number. Any invalid combination of length 
       and loopback will be skipped during the generation/solving process.}
       \tabItem{a negative number in (-1, -{\it bmc\_length}). In this case 
       {\it loopback} is considered a value relative to {\it length}. 
       Any invalid combination of length and loopback will be skipped 
       during the generation/solving process.}
       \tabItem{the symbol '\varvalue{X}', which means ``no loopback'' .}
       \tabItem{the symbol '\varvalue{*}', which means ``all possible loopback from zero to 
       {\it length-1}''.}
       
\opt{ -o \parameter{\filename{\it filename}}}{ {\it filename} is the name of the dumped
dimacs file.  It may contain special symbols which will be
macro-expanded to form the real file name. Possible symbols are: }
       \tabItem{@F: model name with path part.}
       \tabItem{@f: model name without path part.} 
       \tabItem{@k: current problem bound.} 
       \tabItem{@l: current loopback value.} 
       \tabItem{@n: index of the currently processed formula in the property 
       database.} 
       \tabItem{@@: the '@' character.}
\end{cmdOpt}
\end{nusmvCommand}
