% -*-latex-*-
\begin{nusmvCommand} {gen\_ltlspec\_sbmc} {\label{genLtlspecSbmc} 
Dumps into one or more dimacs files the given LTL specification, or
all LTL specifications if no formula is given. Generation and dumping
parameters are the maximum bound and the loopback values.}

\cmdLine{gen\_ltlspec\_sbmc [-h ] | [ -n idx | -p "formula [IN
      context]" | -P "name" ] [-k max\_length] [-l loopback] [-o
    filename]}

This command generates one or more problems, and dumps each problem
into a dimacs file. The BMC encoding used is the one by of Latvala,
Biere, Heljanko and Junttila as described in~\cite{vmcai05}.
%
Each problem is related to a specific problem bound, which
increases from zero ($0$) to the given maximum problem length. Here
\code{max\_length} is the bound of the problem that system is going to
generate and dump.  In this context the maximum problem bound is
represented by the \commandopt{k} command parameter, or by its default
value stored in the environment variable \envvar{bmc\_length}.  The
single generated problem also depends on the \code{loopback} parameter
you can explicitly specify by the \commandopt{l} option, or by its
default value stored in the environment variable
\envvar{bmc\_loopback}.

The property to be used for tghe problem dumping may be specified
using the \commandopt{n idx} or the \commandopt{p "formula"} options.
You may specify dimacs file name by using the option \command{-o
"filename"}, otherwise the default value stored in the environment
variable \envvar{bmc\_dimacs\_filename} will be considered.

\begin{cmdOpt}

\opt{-n \parameter{\natnum{\it index}}}{\natnum{\it index} is the
numeric index of a valid LTL specification formula actually located in
the properties database.}
       
\opt{-p \parameter{"\anyexpr [IN context]"}}{Dumps the \anyexpr specified on
the command-line. \code{context} is the module instance name which
the variables in \anyexpr must be evaluated in.}
            
\opt{-P \parameter{"name"}}{ Checks the LTL property named \filename{name}}

\opt{-k \parameter{\natnum{\it max\_length}}}{\natnum{\it max\_length} is the maximum problem
bound to be generated. Only natural numbers are valid values for this
option. If no value is given the environment variable \envvar{\it
bmc\_length} is considered instead.}

\opt{-l \parameter{\set{\it loopback}{\range{0}{max\_length-1},
       \range{-1}{bmc\_length}, X, *}}}{The {\it loopback} value may be: }
       \tabItem{a natural number in (0, {\it max\_length-1}). A positive sign (`+')
       can be also used as prefix of the number. Any invalid
       combination of length and loopback will be skipped during the
       generation/solving process.}
       \tabItem{a negative number in
       (-1, -{\it bmc\_length}). In this case {\it loopback} is
       considered a value relative to {\it max\_length}.  Any invalid
       combination of length and loopback will be skipped during the
       generation/solving process.}
       \tabItem{the symbol
       `\varvalue{X}', which means ``no loopback".}
       \tabItem{the
       symbol `\varvalue{*}', which means ``all possible loopbacks from
       zero to {\it length-1}" .}

\opt{-o \parameter{\filename{\it filename}}}{\filename{\it filename} is the name of the dumped
dimacs file.  It may contain special symbols which will be
macro-expanded to form the real file name. Possible symbols are: }
       \tabItem{@F: model name with path part. }
       \tabItem{@f: model name without path part.} 
       \tabItem{@k: current problem bound.} 
       \tabItem{@l: current loopback value.} 
       \tabItem{@n: index of the currently processed formula in the property 
       database.} 
       \tabItem{@@: the `@' character.}

\end{cmdOpt}
\end{nusmvCommand}
