% -*-latex-*-
\begin{nusmvCommand}{history} {list previous commands and their event numbers}

\label{History Command}

\cmdLine{history [-h] [<num>]}

Lists previous commands and their event numbers.  This is a \unix-like
history mechanism inside the \nusmv shell.\\

\begin{cmdOpt}
\opt{<num>}{Lists the last \texttt{<num>} events.  Lists the last 30
              events if \texttt{<num>} is not specified. }
\end{cmdOpt}

History Substitution:\\
  The history substitution mechanism is a simpler version of the csh history
  substitution mechanism.  It enables you to reuse words from previously typed
  commands.

The default history substitution character is the `\%' (`!' is default
for shell escapes, and `\#' marks the beginning of a comment). This
can be changed using the \command{set} command. In this description '\%' is
used as the history\_char.  The `\%' can appear anywhere in a line.  A
line containing a history substitution is echoed to the screen after
the substitution takes place.  `\%' can be preceded by a `\' in order
to escape the substitution, for example, to enter a `\%' into an alias
or to set the prompt.

Each valid line typed at the prompt is saved.  If the \command{history}
variable is set (see help page for \command{set}), each line is also echoed
to the history file.  You can use the \command{history} command to list the
previously typed commands. 

Substitutions: \\ At any point in a line these history substitutions
are available.
        
\begin{cmdOpt}
\opt{\%:0}{   Initial word of last command.}

\opt{\%:n}{ n-th argument of last command.}

\opt{\%\$}{   Last argument of last command.}

\opt{\%*}{   All but initial word of last command.}

\opt{\%\%}{    Last command.}

\opt{\%stuf}{ Last command beginning with ``stuf".}

\opt{\%n}{    Repeat the n-th command.}

\opt{\%-n}{  Repeat the n-th previous command.}

\opt{$\widehat{\ }$old$\widehat{\ }$new}{ Replace ``old'' with ``new''
in previous command.  Trailing spaces are significant during
substitution.  Initial spaces are not significant.}
        
\end{cmdOpt}
\end{nusmvCommand}
