% -*-latex-*-
\begin{nusmvCommand} {set} {Sets an environment variable}
 
\cmdLine{set [-h] [<name>] [<value>]}

A variable environment is maintained by the command interpreter.  The
\command{set} command sets a variable to a particular value, and the
\command{unset} command removes the definition of a variable.  If
\command{set} is given no arguments, it prints the current value of
all variables.\\
\begin{cmdOpt}
\opt{<name>}{Variable name}

\opt{<value>} {Value to be assigned to the variable.}

\end{cmdOpt}

Using the \command{set} command to set a variable, without giving any
explicit value is allowed, and sets the variable to \code{1}:\\
\code{\nusmvprompt \command{set} foo}\\
will set the variable foo to \code{1}.\\

Interpolation of variables is allowed when using the \command{set}
command. The variables are referred to with the prefix of '\$'. So for
example, what follows can be done to check the value of a set
variable:\\
\code{\nusmvprompt \command{set} foo bar\\
      \nusmvprompt \command{echo} \$foo\\
             bar\\ }

The last line ``bar" will be the output produced by \nusmv.  Variables
can be extended by using the character `:' to concatenate values. For
example: \\
\code{\nusmvprompt \command{set} foo bar\\
      \nusmvprompt \command{set} foo \$foo:foobar\\
      \nusmvprompt \command{echo} \$foo \\
             bar:foobar\\
}

The variable  \texttt{foo} is extended with the value \texttt{foobar
}.  Whitespace characters may be present within quotes. However,
variable interpolation lays the restriction that the characters ':'
and '/' may not be used within quotes. This is to allow for recursive
interpolation.  So for example, the following is allowed\\
\code{\nusmvprompt \command{set} "foo bar" this\\
      \nusmvprompt \command{echo} \$"foo bar"\\
             this \\
} 

The last line will be the output produced by \nusmv. 
  
But in the following, the  value of the variable  \code{foo/bar } will
not be interpreted correctly:
\code{\nusmvprompt \command{set} "foo/bar" this\\
      \nusmvprompt \command{echo} \$"foo/bar"\\
             foo/bar\\
}

If a variable is not set by the \command{set} command, then the variable is
returned unchanged.  Different commands use environment information
for different purposes.  The command interpreter makes use of the
following parameters:

\begin{cmdOpt}
\opt{autoexec}{ Defines a command string to be automatically executed
after every command processed by the command interpreter.  This is
useful for things like timing commands, or tracing the progress of
optimization.}

\opt{open\_path}{ ``open\_path" (in analogy to the shell-variable PATH)
is a list of colon-separated strings giving directories to be searched
whenever a file is opened for read.  Typically the current directory
(.) is the first item in this list. The standard system library
(typically \stdsyslib) is always implicitly appended to
the current path.  This provides a convenient short-hand mechanism for
reaching standard library files.}

\opt{nusmv\_stderr }{ Standard error (normally \code(stderr)) can be
re-directed to a file by setting the variable \envvar{nusmv\_stderr}.}

\opt{nusmv\_stdout}{ Standard output (normally \code(stdout)) can be
re-directed to a file by setting the variable \envvar{nusmv\_stdout}.}

\end{cmdOpt}
\end{nusmvCommand}
