% -*-latex-*-
\begin{nusmvCommand} {simulate} {Performs a simulation from the current selected state}

\cmdLine{simulate [-h] [-p | -v] [-r | -i [-a]] [-c "constraints" | -t "constraints"] [-k steps]}

Generates a sequence of at most  \code{steps} states (representing a
possible execution of the model), starting from the \command{current
state}. The current state must be set via the \command{pick\_state} or
\command{goto\_state} commands.

It is possible to run the simulation in three ways (according to
different command line policies): deterministic (the default mode),
random and interactive.

The resulting sequence is stored in a trace indexed with an integer
number taking into account the total number of traces stored in the
system. There is a different behavior in the way traces are built,
according to how \emph{current state} is set: \emph{current state} is
always put at the beginning of a new trace (so it will contain at most
steps + 1 states) except when it is the last state of an existent old
trace.  In this case the old trace is lengthened by at most steps
states.\\
\begin{cmdOpt}
\opt{-p}{ Prints current generated trace (only those variables whose
value changed from the previous state).}

\opt{-v}{ Verbosely prints current generated trace (changed and
unchanged state and frozen variables).}

\opt{-r}{ Picks a state from a set of possible future states in a
random way.}

\opt{-i}{ Enables the user to interactively choose every state of the
trace, step by step. If the number of possible states is too high,
then the user has to specify some constraints as simple expression.
These constraints are used only for a single simulation step and are
\emph{forgotten} in the following ones. They are to be intended in an
opposite way with respect to those constraints eventually entered with
the \command{pick\_state} command, or during an interactive simulation session
(when the number of future states to be displayed is too high), that
are \emph{local} only to a single step of the simulation and are
\emph{forgotten} in the next one.

To improve readability of the list of the states which the user must
pick one from, each state is presented in terms of difference with
respect of the previous one.}

\opt{-a}{ Displays all the state and frozen variables (changed and unchanged)
during every step of an interactive session. This option works only if
the \commandopt{i} option has been specified.}

\opt{-c \parameter{"constraints"}}{Performs a simulation in which
  computation is restricted to states satisfying those
  \code{constraints}. The desired sequence of states could not exist
  if such constraints were too strong or it may happen that at some
  point of the simulation a future state satisfying those constraints
  doesn't exist: in that case a trace with a number of states less
  than \code{steps} trace is obtained.  Note: \code{constraints} must
  be enclosed between double quotes \code{" "}.  The expression cannot
  contain \code{next} operators, and is automatically shifted by one
  state in order to constraint only the next steps}

\opt{-t \parameter{"constraints"}}{ Performs a simulation in which
  computation is restricted to states satisfying those
  \code{constraints}. The desired sequence of states could not exist
  if such constraints were too strong or it may happen that at some
  point of the simulation a future state satisfying those constraints
  doesn't exist: in that case a trace with a number of states less
  than \code{steps} trace is obtained.  Note: \code{constraints} must
  be enclosed between double quotes \code{" "}.  The expression can
  contain \code{next} operators, and is NOT automatically shifted by
  one state as done with option \code{-c}}

\opt{\natnum{-k steps}}{ Maximum length of the path according to the
  constraints.  The length of a trace could contain less than
  \code{steps} states: this is the case in which simulation stops in
  an intermediate step because it may not exist any future state
  satisfying those constraints. The default value is determined by the
  \varName{default\_simulation\_steps} environment variable}

\end{cmdOpt}

\end{nusmvCommand}
