% -*-latex-*-
\begin{nusmvCommand} {write\_order} {Writes variable order to file.}

\cmdLine{write\_order [-h] [-b] [(-o | -f) order-file]}

Writes the current order of BDD variables in the file specified via
the \commandopt{o} option. If no option is specified the environment
variable \envvar{output\_order\_file} will be considered. If the
variable \envvar{output\_order\_file} is unset (or set to an empty
value) then standard output will be used.

By default, the bits constituting the scalar variables encoding are
not dumped. When a variable bit should be dumped, the scalar variable
which the bit belongs to is dumped instead if not previously dumped. 
The result is a variable ordering containing only scalar and
boolean model variables.

To dump single bits instead of the corresponding scalar variables,
either the option \commandopt{b} can be specified, or the environment
variable \envvar{write\_order\_dumps\_bits} must be previously set.

When the boolean variable dumping is enabled, the single bits will occur
within the resulting ordering file in the same position that they 
occur at BDD level. 

\begin{cmdOpt}
\opt{-b} {Dumps bits of scalar variables instead of the single
scalar variables. See also the variable \envvar{write\_order\_dumps\_bits}.}

\opt{-o \parameter{\filename{order-file}}} {Sets the environment variable
\envvar{output\_order\_file} to \filename{order-file} and then dumps the
ordering list into that file.}

\opt{-f \parameter{\filename{order-file}}}{ Alias for the
\commandopt{o} option. Supplied for backward compatibility.}

\end{cmdOpt}
\end{nusmvCommand}
