% -*-latex-*-
The main purpose of a model checker is to verify that a model satisfies
a set of desired properties specified by the user.
In \nusmv, the specifications to be checked can be expressed in
two different temporal logics: the Computation Tree Logic CTL, and the
Linear Temporal Logic LTL extended with Past Operators.
CTL and LTL specifications are evaluated by \nusmv in order to
determine their truth or falsity in the FSM. When a specification is
discovered to be false, \nusmv constructs and prints a
counterexample, i.e. a trace of the FSM that falsifies the property.
In this section we will describe model checking of specifications
expressed in CTL, while the next section we consider the case of LTL
specifications.


\section{Computation Tree Logic}
\label{Computation Tree Logic}
\index{Computation Tree Logic}

CTL is a \emph{branching-time} logic: its formulas allow for
specifying properties that take into account the non-deterministic,
branching evolution of a FSM. More precisely, the evolution of a FSM
from a given state can be described as an infinite tree, where the nodes
are the states of the FSM and the branching in due to the
non-determinism in the transition relation. The paths in the tree that
start in a given state are the possible alternative evolutions of the
FSM from that state. In CTL one can express properties that should hold for \emph{all the
paths} that start in a state, as well as for properties that should hold
just for \emph{some of the paths}.

\index{CTL Operators}
Consider for instance CTL formula \code{AF p}. It expresses the
condition that, for \emph{all} the paths (\code{A}) stating from a
state, \emph{eventually in the future} (\code{F}) condition \code{p}
must hold. That is, all the possible evolutions of the system will
eventually reach a state satisfying condition \code{p}.
CTL formula \code{EF p}, on the other hand, requires than there
\emph{exists} some path (\code{E}) that eventually in the future
satisfies \code{p}.

Similarly, formula \code{AG p} requires that condition \code{p} is
always, or \emph{globally}, true in all the states of all the possible
paths, while formula \code{EG p} requires that there is some path along
which condition \code{p} is continuously true.

Other CTL operators are:
\begin{itemize}
\item \code{A [p U q]} and \code{E [p U q]}, 
  requiring condition \code{p} to be true \emph{until} a state is
  reached that satisfies condition \code{q};
\item \code{AX p} and \code{EX p}, 
  requiring that condition \code{p} is true in all or in some of the
  next states reachable from the current state.
\end{itemize}
CTL operators can be nested in an arbitrary way and can be combined
using logic operators (\code{!}, \code{\&}, \code{|}, \code{->},
\code{<->}, \ldots).
Typical examples of CTL formulas are \code{AG ! p} (``condition \code{p}
is absent in all the evolutions''), \code{AG EF p} (``it is always
possible to reach a state where \code{p} holds''), and \code{AG (p -> AF
q)} (``each occurrence of condition \code{p} is followed by an occurrence
of condition \code{q}'').

\index{SPEC keyword}
\index{Specification!CTL}
In \nusmv a CTL specification is given as CTL formula introduced
by the keyword ``\code{SPEC}''.
Whenever a CTL specification is processed, \nusmv checks whether
the CTL formula is true in all the initial states of the model. If this
is not a case, then \nusmv generates a counter-example, that is, a
(finite or infinite) trace that exhibits a valid behavior of the model
that does not satisfy the specification. 
Traces are very useful for identifying the error in the specification that
leads to the wrong behavior.
We remark that the generation of a counter-example trace is not always
possible for CTL specifications. Temporal operators corresponding to
existential path quantifiers cannot be proved false by a showing of a
single execution path.  Similarly, sub-formulas preceded by universal
path quantifier cannot be proved true by a showing of a single execution
path.


\section{Semaphore Example}
\label{CTL Semaphore Example}
\index{Examples!Semaphore!CTL}

Consider the case of the semaphore program described in 
\cref{Examples}.
\index{Safety Property}
A desired property for this program is that it should never be the case
that the two processes \code{proc1} and \code{proc2} are at the same time
in the \code{critical} state (this is an example of a ``safety''
property). This property can be expressed by the following CTL formula:
\begin{alltt}
AG ! (proc1.state = critical & proc2.state = critical)
\end{alltt}
\index{Liveness Property}
Another desired property is that, if \code{proc1} wants to enter its
critical state, it eventually does (this is an example of a ``liveness''
property). This property can be expressed by the following CTL formula:
\begin{alltt}
AG (proc1.state = entering -> AF proc1.state = critical)
\end{alltt}
In order to verify the two formulas on the semaphore model, we add the
two corresponding CTL specification to the program, as follows:
\begin{alltt}
MODULE main
 VAR
   semaphore : boolean;
   proc1     : process user(semaphore);
   proc2     : process user(semaphore);
 ASSIGN
   init(semaphore) := FALSE;
 SPEC AG ! (proc1.state = critical & proc2.state = critical)
 SPEC AG (proc1.state = entering -> AF proc1.state = critical)
MODULE user(semaphore)
 VAR
   state : \{idle, entering, critical, exiting\};
 ASSIGN
   init(state) := idle;
   next(state) :=
     case
       state = idle                  : \{idle, entering\};
       state = entering & !semaphore : critical;
       state = critical              : \{critical, exiting\};
       state = exiting               : idle;
       TRUE                          : state;
     esac;
   next(semaphore) :=
     case
       state = entering : TRUE;
       state = exiting  : FALSE;
       TRUE             : semaphore;
     esac;
 FAIRNESS
   running
\end{alltt}
By running \nusmv with the command
\begin{alltt}
\shellprompt \shelltext{\nusmvtxt semaphore.smv}
\end{alltt}
we obtain the following output:
\begin{alltt}
-- specification AG (!(proc1.state = critical & proc2.state = critical)) 
-- is true
-- specification AG (proc1.state = entering -> AF proc1.state = critical) 
-- is false
-- as demonstrated by the following execution sequence
-> State: 1.1 <-
    semaphore = FALSE
    proc1.state = idle
    proc2.state = idle
-> Input: 1.2 <-
    _process_selector_ = proc1
-- Loop starts here
-> State: 1.2 <-
    proc1.state = entering
-> Input: 1.3 <-
    _process_selector_ = proc2
-> State: 1.3 <-
    proc2.state = entering
-> Input: 1.4 <-
    _process_selector_ = proc2
-> State: 1.4 <-
    semaphore = FALSE
    proc2.state = critical
-> Input: 1.5 <-
    _process_selector_ = proc1
-> State: 1.5 <-
-> Input: 1.6 <-
    _process_selector_ = proc2
-> State 1.6 <-
    proc2.state = exiting
-> Input: 1.7 <-
    _process_selector_ = proc2
-> State 1.7 <-
    semaphore = FALSE
    proc2.state = idle
\end{alltt}
Note that \texttt{\_process\_selector\_} is a special variable which
contains the name of the process that will execute to cause a
transition to the next state. The '\texttt{Input}' section displays
the values of variables that the model has no control over, that is it
cannot change their value. Since processes are chosen
nondeterministically in this model, it has no control over the value
of \texttt{\_process\_selector\_}.\\
\nusmv tells us that the first CTL specification is true:
it is never the case that the two processes will be at the same time in
the critical region.
On the other hand, the second specification is false. \nusmv
produces a counter-example path where initially \code{proc1} goes to
state \code{entering} (state $1.2$), and then a loop starts in which
\code{proc2} repeatedly enters its critical region (state $1.4$) and then
returns to its \code{idle} state (state $1.7$); in the loop, \code{proc1}
is activated only when \code{proc2} is in the critical region (input $1.5$), and is
therefore not able to enter its critical region (state $1.5$). This path
not only shows that the specification is false, it also points out why
can it happen that \code{proc1} never enters its critical region.

Note that in the printout of a cyclic, infinite counter-example the
starting point of the loop is marked by \code{-- loop starts here}.
Moreover, in order to make it easier to follow the action in systems
with a large number of variables, only the values of variables that have
changed in the last step are printed in the states of the trace.
%@node  LTL model checking, Bounded Model Checking, CTL model checking, Tutorial
%  node-name,  next,  previous,  up
